---
title: "note19"
output: html_notebook
---

# 19 FCAT Study: Modeling Attainment-Target Scores

```{r}
library(nlme)
library(lattice)
```

## 19.2 A Fixed-Effects Linear Model Fitted Using the Function lm()

```{r}
data(fcat, package = "nlmeU")
fcat
```

```{r}
opts <- options()
options(contrasts = c("contr.sum", "contr.poly"))
options("cont")
```

```{r}
fm19.1 <-
  lm(scorec ~ id + target, data = fcat)
fm19.1
```

```{r}
options(opts)
```

```{r}
fxd <- coef(fm19.1)
idx <- substr(names(fxd), 1, 2) == "id"
names(fxi <- fxd[idx])
```

```{r}
fxd.id <- c(fxi, "id539" = -sum(fxi))
fxd.id
```

```{r}
idx <- substr(names(fxd), 1, 6) == "target"
names(fxi <- fxd[idx])
(fxd.trgt <- c(fxi, "target9" = -sum(fxi)))
```

## 19.3 A Linear Mixed-Effects Model with Crossed Random Effects Fitted Using the Function lmer()

```{r}
library(lme4)
```

```{r}
system.time(
  fm19.2mer <- lmer(scorec ~ (1 | target) + (1 | id), data = fcat)
)
```

```{r}
fm19.2mer
```

```{r}
summ.merFit <- summary(fm19.2mer)
summ.merFit
```

```{r}
isREML(fm19.2mer)
```

```{r}
cl <- getCall(summ.merFit)
cl
```

```{r}
cl$data
```

```{r}
formula(fm19.2mer)
```

```{r}
fixef(fm19.2mer)
```

```{r}
coef(summ.merFit)
```

```{r}
VCorr <- unlist(VarCorr(fm19.2mer))
VCorr
```

```{r}
sigma(fm19.2mer)
```

```{r}
rnf <- ranef(fm19.2mer)
names(rnf)
```

```{r}
length(plx <- plot(rnf))
```

```{r}
plx[1]
```

```{r}
plx[2]
```

```{r}
dpx <- dotplot(rnf)
dpx[1]
```

```{r}
dpx[2]
```

```{r}
rnf.pVar <- ranef(fm19.2mer, postVar = TRUE)
dpx.pVar <- dotplot(rnf.pVar)
```

```{r}
dpx.pVar[1]
```

```{r}
dpx.pVar[2]
```

```{r}
eVCorr <- sapply(rnf, var)
eVCorr
```

```{r}
VCorr
```

```{r}
all(eVCorr < VCorr)
```

```{r}
rnf.id <- rnf$id
arnf.id <- abs(rnf.id)
afxd.id <- abs(fxd.id)
range(afxd.id - arnf.id)
```

```{r}
rnf.trgt <- rnf$target
arnf.trgt <- abs(rnf.trgt)
afxd.trgt <- abs(fxd.trgt)
range(afxd.trgt - arnf.trgt)
```

```{r}
names(dt <- data.frame(afxd.id, arnf.id))
```

```{r}
names(dt)[2] <- "arnf.id"
```

```{r}
myPanel <- function(x, y, ...) {
  panel.grid(h = -1, v = -1)
  panel.lines(c(0, 3), c(0, 3), lty = 2)
  panel.xyplot(x, y, ...)
}
xyplot(arnf.id ~ afxd.id,
  data = dt, panel = myPanel
)
```

```{r}
detach(package:lme4)
detach(package:Matrix)
rm(rnf, plx)
```

## 19.4 A Linear Mixed-Effects Model with Crossed Random Effects Fitted Using the Function lme()

```{r}
library(nlme)
fcat1 <- within(fcat, one1 <- one2 <- 1L)
fcat1
```

```{r}
system.time(
  fm19.2 <-
    lme(scorec ~ 1,
      random = list(
        one1 = pdIdent(~ target - 1),
        one2 = pdIdent(~ id - 1)
      ),
      data = fcat1
    )
)
```

```{r}
fm19.2
```

```{r}
fm19.2$call$data
```

```{r}
logLik(fm19.2)
```

```{r}
fixef(fm19.2)
```

```{r}
fm19.2$dims$N
```

```{r}
try(getVarCov(fm19.2))
```

```{r}
head(VarCorr(fm19.2))
```

```{r}
intervals(fm19.2)
```

```{r}
rnf <- ranef(fm19.2)
try(plot(rnf))
```

```{r}
rnft <- lapply(rnf, t)
plxLis <- lapply(rnft, qqnorm, plot.it = FALSE)
names(plxLis)
```

```{r}
plx <-
  lapply(plxLis,
    FUN = function(el) xyplot(y ~ x, data = el, grid = TRUE)
  )
```

```{r}
plx[["one1"]]
```

```{r}
plx[["one2"]]
```

```{r}
rsd2 <- resid(fm19.2, type = "pearson")
xyplot(rsd2 ~ target, data = fcat1)
```

## 19.5 A Linear Mixed-Effects Model with Crossed Random Effects and Heteroscedastic Residual Errors Fitted Using lme()

```{r}
nItms <- c(4, 6, 8, 5, 9, 6, 8, 6, 5)
(nms <- levels(fcat1$target))
```

```{r}
names(nItms) <- nms
```

```{r}
fcatm <- within(fcat1, {
  nItems <- nItms[as.numeric(target)]
  scorem <- scorec / nItems
})
fcatm
```

```{r}
varWghts <- 1 / sqrt(nItms)
varWghts
```

```{r}
fxdW <- varWghts[-1] / 0.5
fxdW
```

```{r}
fm19.3 <-
  lme(scorem ~ 1,
    random = list(
      one1 = pdIdent(~ target - 1),
      one2 = pdIdent(~ id - 1)
    ),
    weight = varIdent(form = ~ 1 | target, fixed = fxdW),
    data = fcatm
  )
fm19.3
```

```{r}
summary(fm19.3)$tTable
```

```{r}
VarCorr(fm19.3)
```
