---
title: "note17"
output: html_notebook
---

# 17 PRT Trial: Modeling Muscle Fiber Specific-Force

## 17.2 A Model with Occasion-Specific Random Intercepts for Type-1 Fibers

```{r}
library(nlme)
library(lattice)
data(prt, package = "nlmeU")
prt
```

```{r}
lme.spec.form1 <-
  formula(spec.fo ~ (prt.f + occ.f)^2 + sex.f + age.f + sex.f:age.f + bmi)
prt1 <- subset(prt, fiber.f == "Type 1", select = -fiber.f)
prt1
```

```{r, paged.print=FALSE}
fm17.1 <- lme(lme.spec.form1, random = ~ occ.f - 1 | id, data = prt1)
summary(fm17.1)
```

```{r}
getGroupsFormula(fm17.1)
```

```{r}
str(grpF <- getGroups(fm17.1))
```

```{r}
nF1 <- xtabs(~grpF)
range(nF1)
```

```{r}
nF1[which.min(nF1)]
```

```{r}
str(fm17.1$dims)
```

```{r}
fixed1 <- summary(fm17.1)$tTable
nms <- rownames(fixed1)
nms[7:8] <- c("fLow:fPos", "fMale:FOld")
rownames(fixed1) <- nms
printCoefmat(fixed1, has.Pvalue = TRUE, P.values = TRUE)
```

```{r}
getVarCov(fm17.1)
```

```{r}
VarCorr(fm17.1)
```

```{r}
Ri <- getVarCov(fm17.1, c("5", "275"), type = "conditional")
Ri$"275"
```

```{r}
Ri.5 <- Ri$"5"
dim(Ri.5)
```

```{r}
Ri.5d <- diag(Ri.5)[1:6]
Ri.5d
```

```{r}
sgma <- summary(fm17.1)$sigma
sgma^2
```

```{r}
dt5 <- subset(prt1, select = c(id, occ.f), id == "5")
auxF1 <- function(elv) {
  idx <- 1:min(length(elv), 2)
  elv[idx]
}
i.u5 <- unlist(tapply(rownames(dt5), dt5$occ.f, FUN = auxF1))
i.u5
```

```{r}
dt.u5 <- dt5[i.u5, ]
nms.u5 <- paste(i.u5, dt.u5$occ.f, sep = ".")
nms.u5
```

```{r}
Vi <- getVarCov(fm17.1, "5", type = "marginal")
Vi.5 <- Vi$"5"
Vi.u5 <- Vi.5[i.u5, i.u5]
rownames(Vi.u5) <- nms.u5
Vi.u5
```

```{r}
cov2cor(Vi.u5)
```

```{r}
rnf <- ranef(fm17.1)
vrnf <- var(rnf)
vrnf
```

```{r}
plot(rnf)
```

```{r, fig.width=3, fig.height=3}
library(ellipse)
myPanel <- function(x, y, ...) {
  panel.grid(h = -1, v = -1)
  panel.xyplot(x, y)
  ex1 <- ellipse(getVarCov(fm17.1))
  panel.xyplot(ex1[, 1], ex1[, 2], type = "l", lty = 1)
  ex2 <- ellipse(vrnf)
  panel.xyplot(ex2[, 1], ex2[, 2], type = "l", lty = 2)
}
xyplot(rnf[, 2] ~ rnf[, 1],
  xlab = "Pre-intervention",
  ylab = "Post-intervention",
  xlim = c(-40, 40), ylim = c(-40, 40),
  panel = myPanel
)
```

```{r}
prt1r <- within(prt1, {
  residP1 <- residuals(fm17.1, type = "p")
  fitted1 <- fitted(fm17.1)
})
range(prt1r$residP1)
```

```{r}
xyplot(residP1 ~ fitted1 | occ.f,
  data = prt1r,
  ylim = c(-6, 6),
  type = c("p", "smooth"),
  grid = TRUE
)
```

```{r}
qqnorm(prt1r$residP1)
qqline(prt1r$residP1)
```

## 17.3 A Mean-Variance Model with Occasion-Specific Random Intercepts for Type-1 Fibers

```{r}
fm17.2 <- update(fm17.1,
  weights = varPower(form = ~ fitted(.)),
  data = prt1
)
fm17.2
```

```{r}
intervals(fm17.2)$varStruct
```

```{r}
anova(fm17.1, fm17.2)
```

```{r}
prt2r <- within(prt1, {
  residP2 <- residuals(fm17.2, type = "n")
  fitted2 <- fitted(fm17.2)
})
```

```{r}
xyplot(residP2 ~ fitted2 | occ.f,
  data = prt2r,
  ylim = c(-6, 6),
  type = c("p", "smooth"),
  grid = TRUE
)
```

```{r}
qqnorm(prt2r$residP2)
qqline(prt2r$residP2)
```

## 17.4 Heteroscedastic Fiber-Type×Occasion-Specific Random Intercepts

```{r}
lme.spec.form3 <-
  update(
    lme.spec.form1,
    . ~ . + fiber.f + prt.f:fiber.f + occ.f:fiber.f
  )
fm17.3 <-
  lme(lme.spec.form3,
    random = ~ occ.f:fiber.f - 1 | id,
    data = prt
  )
fm17.3
```

```{r}
fixed.D4 <- summary(fm17.3)$tTable
rnms <- rownames(fixed.D4)
rnms[8:11] <- c("Low:Pos", "Low:Type2", "Pos:Type2", "Male:Old")
rownames(fixed.D4) <- rnms
printCoefmat(fixed.D4, digitis = 3, zap.ind = 5)
```

```{r}
fm17.3cov <- getVarCov(fm17.3, type = "random.effect")
rownames(fm17.3cov)
```

```{r}
nms. <- c("T1.Pre", "T1.Pos", "T2.Pre", "T2.Pos")
dimnames(fm17.3cov) <- list(nms., nms.)
fm17.3cov
```

```{r}
fm17.3cor <- cov2cor(fm17.3cov)
print(fm17.3cor, digits = 2, corr = TRUE, stdevs = FALSE)
```

```{r}
R.5 <- getVarCov(fm17.3, type = "conditional")[["5"]]
dim(R.5)
diag(R.5)[1:6]
```

```{r}
sgma <- fm17.3$sigma
sgma
print(sgma^2)
```

```{r}
CI <- intervals(fm17.3, which = "var-cov")
interv <- CI$reStruct$id
interv
```

```{r}
thDnms <-
  c(
    "sd(T1Pre)", "sd(T1Pos)", "sd(T2Pre)", "sd(T2Pos)",
    "cor(T1Pre,T1Pos)", "cor(T1Pre,T2Pre)", "cor(T1Pre,T2Pos)",
    "cor(T1Pos,T2Pre)", "cor(T1Pos,T2Pos)",
    "cor(T2Pre,T2Pos)"
  )
rownames(interv) <- thDnms
interv
```

```{r, fig.width=5, fig.height=5}
residP3 <- residuals(fm17.3, type = "p")
xyplot(residP3 ~ fitted(fm17.3) | fiber.f:occ.f,
  data = prt,
  type = c("p", "smooth")
)
```

```{r}
qqnorm(residP3)
qqline(residP3)
```

```{r}
Vx <- getVarCov(fm17.3, type = "marginal", individual = "5")
Vmtx.5 <- Vx$"5"
dim(Vmtx.5)
```

```{r}
dt5 <- subset(prt, select = c(id, fiber.f, occ.f), id == "5")
dt5
```

```{r}
i.u5 <- unlist(tapply(rownames(dt5), list(dt5$fiber.f, dt5$occ.f), FUN = auxF1))
i.u5
```

```{r}
dt.u5 <- dt5[i.u5, ]
nms.u5 <- paste(format(i.u5, 2, justify = "right"), abbreviate(dt.u5$fiber.f, 2),
  dt.u5$occ.f,
  sep = "."
)
nms.u5
```

```{r}
Vmtx.u5 <- Vmtx.5[i.u5, i.u5]
dimnames(Vmtx.u5) <- list(nms.u5, i.u5)
Cmtx.u5 <- cov2cor(Vmtx.u5)
uptri <- upper.tri(Cmtx.u5)
Vmtx.u5[uptri] <- Cmtx.u5[uptri]
print(Vmtx.u5, digits = 2)
```

## 17.5 Alternate Form of Heteroscedastic Fiber-Type×Occasion Random Intercepts

```{r}
fm17.3a <-
  lme(lme.spec.form3,
    random = ~ 1 + fiber.f + occ.f + fiber.f:occ.f | id,
    data = prt
  )
fm17.3a
```

```{r}
print(fm17.3a$sigma, digits = 4)
```

```{r}
fm17.3acov <- getVarCov(fm17.3a, type = "random.effect", individual = "5")
dimnames(fm17.3acov)[[1]]
```

```{r}
nms <- c("(Int)", "T2", "Pos", "T2:Pos")
dimnames(fm17.3acov) <- list(nms, nms)
print(fm17.3acov, digits = 4)
```

```{r}
td <- matrix(c(
  1, 0, 0, 0,
  1, 0, 1, 0,
  1, 1, 0, 0,
  1, 1, 1, 1
), nrow = 4, ncol = 4, byrow = TRUE)
mat.D4 <- td %*% fm17.3acov %*% t(td)
dimnames(mat.D4) <- list(nms., nms.)
print(mat.D4, digits = 5)
```

## 17.6 A Model with Heteroscedastic Fiber-Type×Occasion-Specific Random Intercepts and a Structured Matrix D

pdKronecker를 다운받아서 해보았지만, 작동 불가로 더이상 진행하지 않음.

```{r}
# install.packages("nlmeUpdK", repos="http://R-Forge.R-project.org")
library(nlmeUpdK)
```

```{r}
pdId <- pdIdent(~1)
pd1UN <- pdLogChol(~ fiber.f - 1)
pd2UN <- pdLogChol(~ occ.f - 1)
pdL1 <-
  list(
    X = pdId,
    FiberType = pd1UN,
    PrePos = pd2UN
  )

pdKnms <- names(pdL1)
pdKnms
```

```{r}
try(fm17.4 <-
  lme(lme.spec.form3,
    random = list(id = pdKronecker(pdL1)),
    data = prt
  ))
```
