---
title: "Chapter 6"
output: html_notebook
---

# 6 Nonlinear least squares

## 6.1 nls() from package stats

### 6.1.1 A simple example

```{r}
## ----label=C06nlschunk101, echo=TRUE, cache=TRUE, fig.height=5-----------
Weight <- c(
  184.35, 182.51, 180.45, 179.91, 177.91, 175.81, 173.11,
  170.06, 169.31, 165.1, 163.11, 158.3, 155.8, 154.31,
  153.86, 154.2, 152.2, 152.8, 150.3, 147.8, 146.1, 145.6,
  142.5, 142.3, 139.4, 137.9, 133.7, 133.7, 133.3, 131.2,
  133, 132.2, 130.8, 131.3, 129, 127.9, 126.9, 127.7,
  129.5, 128.4, 125.4, 124.9, 124.9, 118.2, 118.2, 115.3,
  115.7, 116, 115.5, 112.6, 114, 112.6
)
Days <- c(
  0, 4, 7, 7, 11, 18, 24, 30, 32, 43, 46, 60, 64, 70, 71,
  71, 73, 74, 84, 88, 95, 102, 106, 109, 115, 122, 133, 137,
  140, 143, 147, 148, 149, 150, 153, 156, 161, 164, 165,
  165, 170, 176, 179, 198, 214, 218, 221, 225, 233, 238,
  241, 246
)
wl <- data.frame(Weight = Weight, Days = Days)
wlmod <- "Weight ~ b0 + b1*2^(-Days/th)"
wlnls <- nls(wlmod, data = wl, start = c(b0 = 1, b1 = 1, th = 1))
summary(wlnls)
```

```{r}
plot(wl$Days, wl$Weight)
title("Weight loss over time in Days")
points(wl$Days, predict(wlnls), type = "l", col = "red")
```

## 6.2 A more difficult case

```{r}
## ----label=C06nlschunk01, echo=TRUE--------------------------------------
# options(width = 60)
pastured <- data.frame(
  time = c(9, 14, 21, 28, 42, 57, 63, 70, 79),
  yield = c(8.93, 10.8, 18.59, 22.33, 39.35, 56.11, 61.73, 64.62, 67.08)
)
plot(yield ~ time, data = pastured)
```

```{r}
regmod <- "yield ~ t1 - t2*exp(-exp(t3+t4*log(time)))"
ones <- c(t3 = 1, t4 = 1, t1 = 1, t2 = 1) # all ones start
anls <- try(nls(regmod, start = ones, trace = FALSE, data = pastured))
anls
```

```{r}
huetstart <- c(t3 = 0, t4 = 1, t1 = 70, t2 = 60)
anlsx <- try(nls(regmod, start = huetstart, trace = FALSE, data = pastured))
anlsx
```

```{r}
library(nlmrt)
anmrt <- nlxb(regmod, start = ones, trace = FALSE, data = pastured)
anmrt
```

```{r}
anmrtx <- try(nlxb(regmod, start = huetstart, trace = FALSE, data = pastured))
anmrtx
```

```{r}
fhard <- function(time, t1, t2, t3, t4) {
  time <- as.vector(time)
  t1 - t2 * exp(-exp(t3 + t4 * log(time)))
}
params <- append(
  list(time = seq(9, 80, length.out = 1001)),
  as.list(anmrtx$coefficients)
)
y <- do.call(fhard, params)
plot(yield ~ time, data = pastured)
lines(params$time, y, col = "red")
```

```{r}
pastjac <- model2jacfun(regmod, ones, funname = "pastjac")
J1 <- pastjac(ones, yield = pastured$yield, time = pastured$time)
svd(J1)$d
## [1] 3.0000e+00 1.3212e-09 2.1637e-16 3.9400e-26
J2 <- pastjac(huetstart, yield = pastured$yield, time = pastured$time)
svd(J2)$d
## [1] 3.0005e+00 1.5143e-01 1.1971e-04 1.8584e-10
```

```{r}
Jnmrtx <- pastjac(coef(anmrtx), yield = pastured$yield, time = pastured$time)
svals <- svd(Jnmrtx)$d
barplot(svals,
  main = "Singular values at nlxb solution to pasture problem",
  horiz = TRUE
)
```

```{r}
require(minpack.lm)
## Loading required package: minpack.lm
aminp <- try(nlsLM(regmod, start = ones, trace = FALSE, data = pastured))
summary(aminp)
```

```{r}
aminpx <- try(nlsLM(regmod, start = huetstart, trace = FALSE, data = pastured))
print(aminpx)
```

```{r}
## ----label=C06nlschunk05b, echo=TRUE-------------------------------------
require(nls2)
set.seed(123) # for reproducibility
regmodf <- as.formula(regmod) # just in case
m100 <- c(t1 = -100, t2 = -100, t3 = -100, t4 = -100)
p100 <- (-1) * m100
gstart <- data.frame(rbind(m100, p100))
anls2 <- try(nls2(regmodf,
  start = gstart, data = pastured,
  algorithm = "random-search", control = list(maxiter = 1000)
))
print(anls2)
```

```{r}
## ----label=C06nlschunk05c, echo=TRUE-------------------------------------
options(show.error.messages = FALSE)
require(nls2)
set.seed(123) # for reproducibility
plinform <- yield ~ cbind(1, -exp(-exp(t3 + t4 * log(time))))
gstartpl <- data.frame(rbind(c(-10, 1), c(10, 8)))
names(gstartpl) <- c("t3", "t4")
anls2plb <- try(nls2(plinform,
  start = gstartpl, data = pastured,
  algorithm = "plinear-brute", control = list(maxiter = 200)
))
options(show.error.messages = TRUE)
print(anls2plb)
```

```{r}
## ====================================
options(show.error.messages = FALSE)
anls2plr <- try(nls2(plinform,
  start = gstartpl, data = pastured,
  algorithm = "plinear-random", control = list(maxiter = 200)
), silent = TRUE)
print(anls2plr)
options(show.error.messages = TRUE)
```

```{r}
# source("./resources/C06pastured1.R", echo = FALSE)
# source("./resources/C06pasturednlxb.R", echo = FALSE)
# source("./resources/C06pasturedSVD.R", echo = FALSE)
# # X11() # ensure new graphics screen
# source("./resources/C06pasturednlsLM.R", echo = FALSE)
# source("./resources/C06pasturednls2.R", echo = FALSE)
#

## ----label=C06nlschunk05d, echo=TRUE, fig.height=4.75--------------------
require(nlmrt)
splb <- coef(anls2plb)
splr <- coef(anls2plr)
# names(splb) <- names(huetstart)
names(splr) <- names(huetstart)
# anlsfromplb <- nls(regmod, start = splb, trace = FALSE, data = pastured)
anlsfromplr <- nls(regmod, start = splr, trace = FALSE, data = pastured)
agomp <- wrapnls(regmodf, start = huetstart, data = pastured)
fitnlxb <- fitted(agomp)
fitnls2b <- fitted(anls2plb)
fitnls2r <- fitted(anls2plr)
plot(pastured$time, fitnlxb, type = "l", lwd = 2, xlab = "time")
points(pastured$time, fitnls2b, col = "red", type = "l", lty = "dashed", lwd = 2)
points(pastured$time, fitnls2r, col = "blue", type = "l", lty = "dotdash", lwd = 2)
points(pastured$time, pastured$yield)
title(main = "Fitted models for pasture data")
sstr <- "best = solid line, nls2.plinear.brute = dashed, nls2.plinear.random = dotdash"
title(sub = sstr, cex.sub = 0.8)
resnlxb <- pastured$yield - fitnlxb
resnls2b <- pastured$yield - fitnls2b
resnls2r <- pastured$yield - fitnls2r


## ----label=C06nlschunk05e, echo=FALSE, cache=TRUE, fig.height=4.75-------
# X11() # added to website file to avoid overwrite of screen plot
# boxplot(data.frame(resnlxb, resnls2b, resnls2r))
title(main = "Model residuals")
title(sub = "best, nls2.plinear.brute, nls2.plinear.random")
```

```{r}
ssfn <- model2ssfun(regmodf, ones)
grfn <- model2grfun(regmodf, ones)
time <- pastured$time
yield <- pastured$yield
# aopto<-optim(ones, ssfn, gr=grfn, yield=yield, time=time,
# method=’BFGS’, control=list(maxit=10000)) print(aopto) ##
# We did not run this one.
```

```{r}
aopto <- optim(ones, ssfn, gr = grfn, yield = yield, time = time, method = "BFGS", control = list(maxit = 10000))
print(aopto)
```

```{r}
aopth <- optim(huetstart, ssfn,
  gr = grfn, yield = yield, time = time,
  method = "BFGS", control = list(maxit = 10000)
)
print(aopth)
```

# 6.3 The structure of the nls() solution

```{r}
require(nlmrt)
wlmod <- "Weight ~ b0 + b1*2^(-Days/th)"
wlnlxb <- nlxb(wlmod, data = wl, start = c(b0 = 1, b1 = 1, th = 1))
wlnlxb
```

```{r}
coef(wlnls)
```

```{r}
print(coef(wlnlxb))
```

## 6.4 Concerns with nls()

### 6.4.1 Small residuals

```{r}
## ----label=C06zeroresprob1, echo=TRUE, cache=TRUE------------------------
require(nlmrt)
x <- 1:10
y <- 2 * x + 3 # perfect fit
yeps <- y + rnorm(length(y), sd = 0.01) # added noise
anoise <- nls(yeps ~ a + b * x, start = list(a = 0.12345, b = 0.54321))
summary(anoise)
```

```{r}
aperf <- try(nls(y ~ a + b * x, start = list(a = 0.12345, b = 0.54321)))
# print(strwrap(aperf))
```

```{r}
ldata <- data.frame(x = x, y = y)
aperfn <- try(nlxb(y ~ a + b * x, start = list(a = 0.12345, b = 0.54321), data = ldata))
aperfn
```

### 6.4.2 Robustness -- "singular gradient" woes

```{r}
x <- c(60, 80, 100, 120)
y <- c(0.8, 6.5, 20.5, 45.9)
mydata <- data.frame(x, y)
pnames <- c("a", "b", "d")
npar <- length(pnames)
st <- c(1, 1, 1)
names(st) <- pnames
plot(x, y)
```

```{r}
rnls <- try(nls(y ~ exp(a + b * x) + d, start = st, data = mydata), silent = TRUE)
if (class(rnls) == "try-error") {
  cat("nls() failed (singular gradient?)\n")
} else {
  summary(rnls)
}
```

```{r}
require(nlmrt)
rnlx0 <- try(nlxb(y ~ exp(a + b * x) + d, start = st, data = mydata), silent = TRUE)
cat("Found sum of squares ", rnlx0$ssquares, "\n")
print(rnlx0$coefficients)
```

```{r}
rnlx0 <- try(nlxb(y ~ exp(a + b * x) + d, start = list(a = 0.5, b = 0.1, d = -5e47), data = mydata), silent = TRUE)
rnlx0
```

```{r}
rnlx1 <- try(nlxb(y ~ exp(a + b * x) + d,
  start = st, data = mydata,
  control = list(rofftest = FALSE)
), silent = TRUE)
# cat("Found sum of squares ",rnlx1$ssquares,"\n")
rnlx1
```

```{r}
rnlx2 <- try(nlxb(y ~ exp(a + b * x) + d,
  start = st, data = mydata,
  control = list(rofftest = FALSE, smallsstest = FALSE)
), silent = TRUE)
# cat("Found sum of squares ",rnlx2$ssquares,"\n")
rnlx2
```

### Bounds with nls()

```{r}
## ----label=C06boundnls1, echo=TRUE, cache=TRUE---------------------------
ydat <- c(
  5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
  38.558, 50.156, 62.948, 75.995, 91.972
) # for testing
tdat <- 1:length(ydat) # for testing
weeddata <- data.frame(y = ydat, t = tdat)
plot(tdat, ydat)
```

```{r}
require(nlmrt)
hobmod <- "y~100*b1/(1+10*b2*exp(-0.1*b3*t))"
st <- c(b1 = 1, b2 = 1, b3 = 1)
low <- c(-Inf, -Inf, -Inf)
up <- c(2, 2, 2)
# up <- c(6, 6, 6)

cat("try nlxb\n")
anlxb2 <- nlxb(hobmod, st, data = weeddata, lower = low, upper = up)
anlxb2
```

```{r}
try(anls2p <- nls(hobmod, st, data = weeddata, lower = low, upper = up, algorithm = "port"))
summary(anls2p)
anls2p$m$deviance()
```

```{r}
require(minpack.lm)
aLM2 <- nlsLM(hobmod, st, data = weeddata, lower = low, upper = up)
summary(aLM2)
aLM2$m$deviance()
```

## 6.5 Some ancillary tools for nonlinear least squares

### 6.5.1 Starting values and self-starting problems

```{r}
## ----label=C06hobbselfstart2, echo=TRUE----------------------------------
## Put in weeddata here as precaution. Maybe reset workspace.
anlss2 <- nls(y ~ SSlogis(t, p1, p2, p3), data = weeddata)
summary(anlss2)
anlss2$m$deviance()
```

## 6.6 Minimizing R functions that compute sums of squares

```{r}
## ----label=C06wtdnlls1, echo=TRUE, cache=TRUE----------------------------
## weighted nonlinear regression
Treated <- as.data.frame(Puromycin[Puromycin$state == "treated", ])
plot(rate ~ conc, data = Treated)
```

```{r}
weighted.MM <- function(resp, conc, Vm, K) {
  ## Purpose: exactly as white book p. 451 -- RHS for nls()
  ##  Weighted version of Michaelis-Menten model
  ## ----------------------------------------------------------
  ## Arguments: 'y', 'x' and the two parameters (see book)
  ## ----------------------------------------------------------
  ## Author: Martin Maechler, Date: 23 Mar 2001

  pred <- (Vm * conc) / (K + conc)
  (resp - pred) / sqrt(pred)
}

start <- list(Vm = 200, K = 0.1)
Pur.wt <- nls(~ weighted.MM(rate, conc, Vm, K),
  data = Treated,
  start
)
print(summary(Pur.wt))
```

```{r}
wtMM <- function(x, resp, conc) { # redefined for nlfb()
  Vm <- x[1]
  K <- x[2]
  res <- weighted.MM(resp, conc, Vm, K)
}
anlf <- nlfb(start,
  resfn = wtMM, jacfn = NULL, trace = FALSE,
  conc = Treated$conc, resp = Treated$rate
)
anlf
```

## 6.7 Choosing an approach

```{r}
## ----label=C06tsl1, echo=TRUE, cache=TRUE, fig.height=4------------------
set.seed(1235)
x <- 1:40
xint <- 20.5 * rep(1, length(x))
sla <- 0.5
slb <- -0.5
yint <- 30
idx <- which(x <= xint)
ymod <- {
  yint + (x - xint) * slb
}
ymod[idx] <- yint + (x[idx] - xint[idx]) * sla
ydata <- ymod + rnorm(length(x), 0, 2)
plot(x, ymod, type = "l", ylim = c(0, 40), ylab = "y")
points(x, ydata)
title(main = "2 straight lines data")
title(sub = "Lines are those used to generate data")
```

```{r}
tslexp <- "y ~ (yint + (x-xint)*slb)*(x >= xint) + (yint + (x-xint)*sla)*( x < xint)"
mydf <- data.frame(x = x, y = ydata)
mnls <- try(nls(tslexp, trace = FALSE, data = mydf))
strwrap(mnls)
```

```{r}
mystart <- c(xint = 1, yint = 1, sla = 1, slb = 1)
mnls <- try(nls(tslexp, start = mystart, trace = FALSE, data = mydf))
strwrap(mnls)
```

```{r}
myst2 <- c(xint = 15, yint = 25, sla = 1, slb = -1)
mnls2 <- try(nls(tslexp, start = myst2, trace = FALSE, data = mydf))
summary(mnls2)
mnls2$m$deviance()
```

```{r}
mnlxb <- try(nlxb(tslexp, start = mystart, trace = FALSE, data = mydf))
strwrap(mnlxb)
```

```{r}
myres <- model2resfun(tslexp, mystart)
tres <- myres(myst2, x = mydf$x, y = mydf$y)
tres
```

```{r}
jactry <- try(myjac <- model2jacfun(tslexp, mystart))
strwrap(jactry)
```

```{r}
## Try nlfb with no jacfn.
mnlfb <- try(nlfb(mystart, resfn = myres, trace = FALSE, x = mydf$x, y = mydf$y))
mnlfb
```

```{r}
## Also try from better start
mnlfb2 <- try(nlfb(myst2, resfn = myres, trace = FALSE, x = mydf$x, y = mydf$y))
mnlfb2
```

```{r}
## It will also work with
## mnlfb2j<-try(nlfb(myst2, resfn=myres, jacfn=NULL, trace=FALSE, x=mydf$x, y=mydf$y))
## or with Package pracma's function lsqnonlin (but there is not a good summary())
require(pracma)
alsqn <- lsqnonlin(myres, x0 = myst2, x = mydf$x, y = mydf$y)
print(alsqn$ssq)
print(alsqn$x)
```

## 6.8 Separable sums of squares problems

```{r}
require(NISTnls)
L1 <- Lanczos1 # The data frame of data, but more than 2 decimals on y
L1[, "y"] <- round(L1[, "y"], 2)
L1
plot(y ~ x, data = L1)
```

```{r}
## ----label=C06lanc2, echo=TRUE, cache=TRUE-------------------------------
lanc.res <- function(b, data) { # 3 exponentials
  res <- rep(NA, length(data$x))
  res <- b[4] * exp(-b[1] * data$x) + b[5] * exp(-b[2] * data$x) +
    b[6] * exp(-b[3] * data$x) - data$y
}
lanc.jac <- function(b, data) { # 3 exponentials
  expr3 <- exp(-b[1] * data$x)
  expr7 <- exp(-b[2] * data$x)
  expr12 <- exp(-b[3] * data$x)
  J <- matrix(0, nrow = length(data$x), ncol = length(b))
  J[, 4] <- expr3
  J[, 1] <- -(b[4] * (expr3 * data$x))
  J[, 5] <- expr7
  J[, 2] <- -(b[5] * (expr7 * data$x))
  J[, 6] <- expr12
  J[, 3] <- -(b[6] * (expr12 * data$x))
  J
}
```

```{r}
bb1 <- c(1, 2, 3, 4, 5, 6)
bb2 <- c(1, 1.1, 1.2, .1, .1, .1)
names(bb1) <- c("a1", "a2", "a3", "c1", "c2", "c3")
names(bb2) <- c("a1", "a2", "a3", "c1", "c2", "c3")
## Check residual function
cat("Sumsquares at bb1 =", as.numeric(crossprod(lanc.res(bb1, data = L1))))
require(numDeriv)
JJ <- lanc.jac(bb1, data = L1)
JJn <- jacobian(lanc.res, bb1, data = L1)
cat("max abs deviation JJ and JJn:", max(abs(JJ - JJn)), "\n")
lancexp <- "y ~ (c1*exp(-a1*x) + c2*exp(-a2*x) + c3*exp(-a3*x))"
```

```{r}
## ----label=C06lanc3, echo=TRUE, cache=TRUE-------------------------------
tnls1 <- try(nls(lancexp, start = bb1, data = Lanczos1, trace = FALSE))
strwrap(tnls1)
```

```{r}
tnls2 <- try(nls(lancexp, start = bb2, data = Lanczos1, trace = FALSE))
strwrap(tnls2)
```

```{r}
## nlmrt
require(nlmrt)
tnlxb1 <- try(nlxb(lancexp, start = bb1, data = Lanczos1, trace = FALSE))
tnlxb1
tnlxb2 <- try(nlxb(lancexp, start = bb2, data = Lanczos1, trace = FALSE))
tnlxb2
tnlfb1 <- try(nlfb(bb1, lanc.res, trace = FALSE, data = Lanczos1))
tnlfb1
tnlfb2 <- try(nlfb(bb2, lanc.res, trace = FALSE, data = Lanczos1))
tnlfb2
tnlfb1j <- try(nlfb(bb1, lanc.res, lanc.jac, trace = FALSE, data = Lanczos1))
tnlfb1j
tnlfb2j <- try(nlfb(bb2, lanc.res, lanc.jac, trace = FALSE, data = Lanczos1))
tnlfb2j
```

```{r}
bb1s <- bb1[1:3] # shorten start, as only 3 starting nonlinear parameters
bb2s <- bb2[1:3]
llexp <- "y~cbind(exp(-a1*x), exp(-a2*x), exp(-a3*x))"
tnls1p <- try(nls(llexp,
  start = bb1s, data = Lanczos1, algorithm = "plinear",
  trace = TRUE, control = list(maxiter = 100)
))
strwrap(tnls1p)
```

```{r}
tnls2p <- try(nls(llexp,
  start = bb2s, data = Lanczos1, algorithm = "plinear",
  trace = TRUE, control = list(maxiter = 100)
))
strwrap(tnls2p)
```

```{r}

## ----label=C06lanc5, echo=TRUE, cache=TRUE-------------------------------
lanclin.res <- function(b, data) { # restructured to allow for easier linearization
  xx <- data$x
  yy <- data$y
  res <- rep(NA, length(xx))
  m <- length(xx)
  n <- 3
  A <- matrix(NA, nrow = m, ncol = n)
  for (j in 1:n) {
    A[, j] <- exp(-b[j] * xx)
  }
  lmod <- lsfit(A, yy, intercept = FALSE)
  res <- lmod$residuals
  attr(res, "coef") <- lmod$coef
  res
}
bb1s
res1L <- lanclin.res(bb1s, data = Lanczos1)
cat(
  "sumsquares via lanclin for bb1s start:",
  as.numeric(crossprod(res1L)), "\n"
)
tnlfbL1 <- try(nlfb(bb1s, lanclin.res, trace = FALSE, data = Lanczos1))
tnlfbL1
## Get the linear coefficients
resL1 <- lanclin.res(coef(tnlfbL1), data = Lanczos1)
resL1
```

```{r}
## Now from bb2s
bb2s
res2L <- lanclin.res(bb2s, data = Lanczos1)
cat(
  "sumsquares via lanclin for bb2s start:",
  as.numeric(crossprod(res2L)), "\n"
)
tnlfbL2 <- try(nlfb(bb2s, lanclin.res, trace = FALSE, data = Lanczos1))
tnlfbL2
## Get the linear coefficients
resL2 <- lanclin.res(coef(tnlfbL2), data = Lanczos1)
resL2
```

## Lanczos1 help

```{r}
Try <- function(expr) if (!inherits(val <- try(expr), "try-error")) val
plot(y ~ x, data = Lanczos1)
## plot on log scale to see the apparent number of exponential terms
plot(y ~ x, data = Lanczos1, log = "y")
```

```{r}
## data are an exact fit so the convergence criterion fails
Try(fm1 <- nls(y ~ b1 * exp(-b2 * x) + b3 * exp(-b4 * x) + b5 * exp(-b6 * x),
  data = Lanczos1, trace = TRUE,
  start = c(
    b1 = 1.2, b2 = 0.3, b3 = 5.6, b4 = 5.5,
    b5 = 6.5, b6 = 7.6
  )
))
```

```{r}
Try(fm1a <- nls(y ~ b1 * exp(-b2 * x) + b3 * exp(-b4 * x) + b5 * exp(-b6 * x),
  data = Lanczos1, trace = TRUE, alg = "port",
  start = c(
    b1 = 1.2, b2 = 0.3, b3 = 5.6,
    b4 = 5.5, b5 = 6.5, b6 = 7.6
  )
))
```

```{r}
## data are an exact fit so the convergence criterion fails
Try(fm2 <- nls(y ~ b1 * exp(-b2 * x) + b3 * exp(-b4 * x) + b5 * exp(-b6 * x),
  data = Lanczos1, trace = TRUE,
  start = c(
    b1 = 0.5, b2 = 0.7, b3 = 3.6,
    b4 = 4.2, b5 = 4, b6 = 6.3
  )
))
```

```{r}
Try(fm2a <- nls(y ~ b1 * exp(-b2 * x) + b3 * exp(-b4 * x) + b5 * exp(-b6 * x),
  data = Lanczos1, trace = TRUE, alg = "port",
  start = c(
    b1 = 0.5, b2 = 0.7, b3 = 3.6,
    b4 = 4.2, b5 = 4, b6 = 6.3
  )
))
```

```{r}
## data are an exact fit so the convergence criterion fails
Try(fm3 <- nls(y ~ exp(outer(x, -c(b2, b4, b6))),
  data = Lanczos1, trace = TRUE, algorithm = "plinear",
  start = c(b2 = 0.3, b4 = 5.5, b6 = 7.6)
))
```
