---
title: "note18"
output: html_notebook
---

# 18 SII Project: Modeling Gains in Mathematics Achievement-Scores

## 18.2 A Model with Fixed Effects for School- and Pupil-Specific Covariates and Random Intercepts for Schools and Classes

```{r}
library(nlme)
library(lattice)
data(SIIdata, package = "nlmeU")
```

```{r}
form1 <-
  formula(mathgain ~ ses + minority + mathkind + sex + housepov)
fm18.1 <-
  lme(form1,
    random = list(schoolid = ~1, classid = ~1),
    data = SIIdata,
    method = "ML"
  )
fm18.1
```

```{r}
update(fm18.1,
  random = ~ 1 | schoolid / classid
)
```

```{r}
getGroupsFormula(fm18.1)
```

```{r}
str(grpF1 <- getGroups(fm18.1, level = 1))
```

```{r}
str(grpF2 <- getGroups(fm18.1))
```

```{r}
grpF2
```

```{r}
fxd <- fixef(fm18.1)
fxd
```

```{r}
vcov1 <- vcov(fm18.1)
nms <- abbreviate(names(fxd))
dimnames(vcov1) <- list(nms, nms)
print(vcov1, digits = 2)
```

```{r}
try(getVarCov(fm18.1))
```

```{r}
VarCorr(fm18.1)
```

```{r}
anova(fm18.1, type = "marginal")
```

```{r}
anova(fm18.1, Terms = c("housepov"))
```

```{r}
anova(fm18.1, Terms = c("sex"))
```

```{r}
try(anova(fm18.1, Terms = c("housepov", "sex")))
```

```{r}
rsd1 <- resid(fm18.1, level = 0)
range(rsd1)
```

```{r}
outi <- abs(rsd1) > 120
as.numeric(SIIdata$childid[outi])
rsd1[outi]
```

```{r}
myPanel <- function(x, y, subscripts, ...) {
  panel.xyplot(x, y, ...)
  outi <- abs(y) > 120
  y1 <- y[outi]
  x1 <- x[outi]
  ltext(x1, y1, names(y1), pos = 3)
}
xyplot(rsd1 ~ housepov | sex, SIIdata,
  type = c("p", "r"),
  panel = myPanel
)
```

## 18.3 A Model with an Interaction Between School- and Pupil-Level Covariates

```{r, paged.print=FALSE}
form2 <- update(form1, . ~ . + sex:housepov)
fm18.2 <- update(fm18.1, form2)
summary(fm18.2)
```

```{r}
anova(fm18.2, Terms = "sex:housepov")
```

```{r}
anova(fm18.1, fm18.2)
```

## 18.4 A Model with Fixed Effects of Pupil-Level Covariates Only

```{r}
form3 <- update(form1, . ~ . - sex - housepov)
fm18.3 <- update(fm18.1, form3)
anova(fm18.1, fm18.3, fm18.2)
```

```{r}
xyplot(resid(fm18.3) ~ mathkind, SIIdata, type = c("p", "smooth"))
```

```{r}
xyplot(resid(fm18.3) ~ ses, SIIdata, type = c("p", "smooth"))
```

## 18.5 A Model with a Third-Degree Polynomial of a Pupil-Level Covariate in the Mean Structure

```{r}
form4 <-
  formula(mathgain ~ ses + minority + poly(mathkind, 3))
fm18.4 <- update(fm18.3, form4)
fm18.4
```

```{r}
anova(fm18.3, fm18.4)
```

```{r}
summary(fm18.4)
```

```{r}
auxL <- list(
  ses = 0,
  minority = factor(c("Mnrt=No", "Mnrt=Yes")),
  mathkind = seq(290, 625, by = 5)
)
auxDt <- expand.grid(auxL)
auxDt
```

```{r}
prd <- predict(fm18.4, auxDt, level = 0)
prd4Dt <- data.frame(auxDt, pred4 = prd)
prd4Dt
```

```{r}
xyplot(pred4 ~ mathkind,
  groups = minority, data = prd4Dt,
  type = "l", grid = TRUE, ylim = c(-70, 220)
)
```

```{r}
auxL <- list(
  ses = seq(-1.5, 3.5, by = 0.1),
  minority = factor(c("Mnrt=No", "Mnrt=Yes")),
  mathkind = 450
)
auxDt <- expand.grid(auxL)
auxDt

prd <- predict(fm18.4, auxDt, level = 0)
prd4Dt <- data.frame(auxDt, pred4 = prd)
prd4Dt

xyplot(pred4 ~ ses,
  groups = minority, data = prd4Dt,
  type = "l", grid = TRUE, ylim = c(-70, 220)
)
```

```{r}
par(mfrow = c(2, 1))
p1 <- xyplot(resid(fm18.4) ~ mathkind, SIIdata, type = c("p", "smooth"), grid = TRUE, cex = 0.3)
p2 <- xyplot(resid(fm18.4) ~ ses, SIIdata, type = c("p", "smooth"), grid = TRUE, cex = 0.3)
print(p1, split = c(1, 1, 2, 1), more = TRUE)
print(p2, split = c(2, 1, 2, 1))
```

## 18.6 A Model with a Spline of a Pupil-Level Covariate in the Mean Structure

```{r, paged.print=FALSE}
require(splines)
form5 <-
  formula(mathgain ~ ses + minority + bs(mathkind, df = 4))
fm18.5 <- update(fm18.4, form5)
summary(fm18.5)
```

```{r, paged.print=TRUE}
AIC(fm18.3, fm18.4, fm18.5)
```

## 18.7 The Final Model with Only Pupil-Level Variables in the Mean Structure

```{r}
form6 <-
  formula(mathgain ~ ses + minority + poly(mathkind, 3) + ses:minority)
fm18.6 <- update(fm18.4, form6)
fm18.6
```

```{r}
anova(fm18.4, fm18.6)
```

```{r}
rsd6 <- resid(fm18.6, level = 0)
xyplot(rsd6 ~ ses | minority, SIIdata, type = c("p", "smooth"))
```

```{r}
qqnorm(fm18.6)
qqnorm(fm18.6, form = ~ resid(., type = "p", level = 2))
qqnorm(fm18.6, form = ~ resid(., type = "p", level = 1))
qqnorm(fm18.6, form = ~ resid(., type = "p", level = 0))
```

```{r}
qqnorm(fm18.6, form = ~ resid(., type = "p") | sex * minority, id = 0.0005)
```

```{r}
rsd6p <- resid(fm18.6, type = "p")
keep <- abs(rsd6p) < 3
rsd6x <- rsd6p[keep]
rsd6p[!keep]
```

```{r}
qqDtx <- qqnorm(rsd6x, plot.it = FALSE)
xp1 <- xyplot(x ~ y, data.frame(qqDtx))
update(xp1,
  xlab = "Qauntiles of standard normal",
  ylab = "Standarized residuals",
  grid = TRUE
)
```

```{r}
qqDtx2 <- cbind(SIIdata[keep, ], qqDtx)
xp2 <- xyplot(x ~ y | sex * minority, data = data.frame(qqDtx2))
update(xp2,
  xlab = "Qauntiles of standard normal",
  ylab = "Standarized residuals",
  grid = TRUE
)
```

```{r}
ref6 <- ranef(fm18.6)
str(ref6)
```

```{r}
pref6 <- plot(ref6)
pref6
```

```{r}
pref6lims <- pref6$y.limits
len <- length(pref6lims)
sel <- seq(1, len, by = 15)
pref6lims[-sel] <- ""
update(pref6, ylim = pref6lims, ylab = "classid %in% schoolid")
```

```{r}
ref61 <- ranef(fm18.6, level = 1)
plot(ref61)
```

```{r}
qqnorm(fm18.6, ~ ranef(., level = 2), id = 0.2, xlab = "Random effects for classes")
```

```{r}
qqnorm(fm18.6, ~ ranef(., level = 1), id = 0.2, xlab = "Random effects for schools")
```

## 18.8 Analysis Using the Function lmer()

```{r}
library(lme4)
```

```{r}
fm18.6mer <-
  lmer(mathgain ~ ses + minority + poly(mathkind, 3) + ses:minority + (1 | schoolid) + (1 | schoolid:classid),
    data = SIIdata,
    REML = FALSE
  )
summ <- summary(fm18.6mer)
```

```{r}
print(summ, corr = FALSE)
```

```{r}
update(
  fm18.6mer,
  mathgain ~ ses + minority + poly(mathkind, 3) + ses:minority +
    (1 | schoolid) + (1 | classid)
) |>
  summary()
```

```{r}
update(
  fm18.6mer,
  mathgain ~ ses + minority + poly(mathkind, 3) + ses:minority +
    (1 | schoolid / classid)
) |>
  summary()
```

```{r}
anova(fm18.6mer)
```

```{r}
logLik(fm18.6mer)
```

```{r}
unlist(VarCorr(fm18.6mer))
```

```{r}
sigma(fm18.6mer)
```

```{r}
rsd6 <- resid(fm18.6mer)
qqnorm(rsd6)
```

```{r}
rnf6mer <- ranef(fm18.6mer)
rnf6qn <- plot(rnf6mer, grid = TRUE)
update(rnf6qn[["schoolid:classid"]], ylab = c("Random effects for classes"))
```

```{r}
update(rnf6qn[["schoolid"]], ylab = c("Random effects for classes"))
```
