---
title: "note16"
output: html_notebook
---

# 16 ARMD Trial: Modeling Visual Acuity

## 16.2 A Model with Random Intercepts and Homogeneous Residual Variance

```{r}
library(nlme)
library(lattice)
data(armd, package = "nlmeU")
```

```{r}
lm2.form <-
  formula(visual ~ visual0 + time + treat.f + treat.f:time)
fm16.1 <-
  lme(lm2.form,
    random = ~ 1 | subject,
    data = armd
  )
fm16.1
```

```{r}
printCoefmat(summary(fm16.1)$tTable, has.Pvalue = TRUE, P.values = TRUE)
```

```{r}
getGroupsFormula(fm16.1)
```

```{r}
str(grpF <- getGroups(fm16.1))
```

```{r}
grpF[1:17]
```

```{r}
levels(grpF)[1:5]
```

```{r}
range(xtabs(~grpF))
```

```{r}
getVarCov(fm16.1, individuals = "1")
```

```{r}
VarCorr(fm16.1)
```

```{r}
getVarCov(fm16.1, type = "conditional", "2")
```

```{r}
fm16.1cov <- getVarCov(fm16.1, type = "marginal", individual = "2")
fm16.1cov
```

```{r}
cov2cor(fm16.1cov[[1]])
```

## 16.3 A Model with Random Intercepts and the varPower(·) Residual Variance-Function

```{r}
fm16.2 <-
  update(fm16.1,
    weights = varPower(form = ~time),
    data = armd
  )
fm16.2
```

```{r}
anova(fm16.1, fm16.2)
```

```{r}
VarCorr(fm16.2)
```

```{r}
getVarCov(fm16.2, type = "conditional", individual = "2")
```

```{r}
fm16.2cov <- getVarCov(fm16.2, type = "marginal", individual = "2")
fm16.2cov
```

```{r}
cov2cor(fm16.2cov[[1]])
```

```{r}
plot(fm16.2)
```

```{r}
bwplot(resid(fm16.2, type = "p") ~ time.f | treat.f,
  # panel = panel.bwxplot2
  data = armd
)
```

```{r}
qqnorm(fm16.2, ~ resid(.) | time.f)
```

```{r}
qqnorm(fm16.2, ~ ranef(.))
```

```{r}
id <- 0.05
outliers.idx <-
  within(armd, {
    resid.p <- resid(fm16.2, type = "pearson")
    idx <- abs(resid.p) > -qnorm(id / 2)
  })
outliers <- subset(outliers.idx, idx)
outliers
```

```{r}
nrow(outliers)
```

```{r}
outliers$subject
```

```{r, fig.width=3, fig.height=3}
aug.Pred <-
  augPred(fm16.2,
    primary = ~time,
    level = 0:1,
    length.out = 2
  )
plot(aug.Pred,
  layout = c(4, 4, 2),
  key = list(
    lines = list(lty = c(1, 2)),
    text = list(c("Marginal", "Subject-specific")),
    columns = 2
  )
)
```

## 16.4 Models with Random Intercepts and Slopes and the varPower(·) Residual Variance-Function

```{r}
fm16.3 <-
  update(fm16.2,
    random = ~ 1 + time | subject,
    data = armd
  )
fm16.3
```

```{r}
getVarCov(fm16.3, individual = "2")
```

```{r}
intervals(fm16.3, which = "var-cov")
```

```{r}
anova(fm16.1, fm16.2, fm16.3)
```

```{r}
fm16.4 <-
  update(fm16.3,
    random = list(subject = pdDiag(~time)),
    data = armd
  )
fm16.4
```

```{r}
intervals(fm16.4)
```

```{r}
anova(fm16.1, fm16.2, fm16.3, fm16.4)
```

```{r}
bwplot(resid(fm16.4, type = "p") ~ time.f | treat.f,
  # panel = panel.bwxplot2
  data = armd
)
```

```{r}
qqnorm(fm16.4, ~ resid(.) | time.f)
```

```{r}
qqnorm(fm16.4, ~ ranef(.))
```

```{r, fig.width=3, fig.height=3}
aug.Pred <-
  augPred(fm16.4,
    primary = ~time,
    level = 0:1,
    length.out = 2
  )
plot(aug.Pred,
  layout = c(4, 4, 1),
  key = list( # lines = list(lty = c(1, 1), col = c(3, 4)),
    text = list(c("Marginal", "Subject-specific")),
    columns = 2
  )
)
```

```{r}
lm3.form <- formula(visual ~ visual0 + time + treat.f)
fm16.5 <-
  update(fm16.4,
    lm3.form,
    data = armd
  )
fm16.5
```

```{r}
printCoefmat(summary(fm16.5)$tTable, has.Pvalue = TRUE, P.values = TRUE)
```

```{r}
intervals(fm16.5, which = "var-cov")
```

```{r}
VarCorr(fm16.5)
```

```{r}
getVarCov(fm16.5, type = "conditional", individual = "2")
```

```{r}
fm16.5cov <- getVarCov(fm16.5, type = "marginal", individual = "2")
fm16.5cov
```

```{r}
cov2cor(fm16.5cov[[1]])
```

```{r}
anova(fm16.4, fm16.5)
```

## 16.5 An Alternative Residual Variance Function: varIdent(·)

```{r}
fm16.6 <-
  update(fm16.3,
    weights = varIdent(form = ~ 1 | time.f)
  )
fm16.6
```

```{r}
try(intervals(fm16.6, ))
```

```{r}
qqnorm(fm16.6, ~ resid(.) | time.f)
```

```{r}
anova(fm16.3, fm16.6)
```

## 16.6 Testing Hypotheses About Random Effects

```{r}
AIC(fm16.1, fm16.2, fm16.3, fm16.4)
```

```{r}
fm16.4ml <- update(fm16.4, method = "ML")
fm16.5ml <- update(fm16.5, method = "ML")
anova(fm16.4ml, fm16.5ml)
```

```{r}
vis.gls1a <- gls(lm2.form, data = armd)
vis.gls1a
```

```{r}
anova.res <- anova(vis.gls1a, fm16.1)
anova.res
```

```{r}
(anova.res[["p-value"]][2]) / 2
```

```{r}
library(RLRsim)
exactRLRT(fm16.1)
```

```{r}
fm16.7 <-
  update(fm16.4,
    weights = NULL,
    data = armd
  )
fm16.7
```

```{r}
an.res <- anova(fm16.1, fm16.7)
an.res
```

```{r}
RLRT <- an.res[["L.Ratio"]][2]
RLRT
```

```{r}
.5 * pchisq(RLRT, 1, lower.tail = FALSE) + .5 * pchisq(RLRT, 2, lower.tail = FALSE)
```

```{r}
mAux <-
  update(fm16.1,
    random = ~ 0 + time | subject,
    data = armd
  )
mAux
```

```{r}
exactRLRT(m = mAux, m0 = fm16.1, mA = fm16.7)
```

```{r}
vis.lme2.sim <-
  simulate(fm16.1, m2 = fm16.7, nsim = 100) # nsim = 10000 takes ~5 min
plot(vis.lme2.sim, df = c(1, 2), abline = c(0, 1, lty = 2))
```

## 16.7 Analysis Using the Function lmer()

```{r}
library(lme4)
fm16.1mer <-
  lmer(visual ~ visual0 + time * treat.f + (1 | subject),
    data = armd
  )
fm16.1mer
# print(fm16.1mer, corr = FALSE)
```

```{r}
vcovb <- vcov(fm16.1mer)
corb <- cov2cor(vcovb)
nms <- abbreviate(names(fixef(fm16.1mer)), 5)
rownames(corb) <- nms
corb
```

```{r}
VarCorr(fm16.1mer)
```

```{r}
sgma <- sigma(fm16.1mer)
sgma
```

```{r}
A <- getME(fm16.1mer, "A")
I.n <- Diagonal(ncol(A))
V <- sgma^2 * (I.n + crossprod(A))
str(getME(fm16.1mer, "flist"))
V
```

```{r}
coefs <- coef(summary(fm16.1mer))
ddf <- c(631, 231, 631, 231, 631)
pT <- 2 * (1 - pt(abs(coefs[, "t value"]), ddf))
tTable <- cbind(coefs, ddf, pT)
printCoefmat(tTable, P.avlues = TRUE, has.Pvalue = TRUE)
```

```{r}
dtaov <- anova(fm16.1mer)
dtaov
```

```{r}
ddf1 <- ddf[-1]
within(dtaov, {
  `Pr(>F)` <- pf(`F value`, npar, ddf1, lower.tail = FALSE)
  denDf <- ddf1
})
```

```{r}
merObject <- fm16.1mer
simD1 <- simulate(merObject, nsim = 1000)
SimD1summ <- apply(
  simD1, 2,
  function(y) {
    auxFit <- refit(merObject, y)
    # summ <- summary(auxFit)
    beta <- fixef(auxFit)
    Sx <- getME(auxFit, "theta")
    sgma <- sigma(auxFit)
    list(beta = beta, St = Sx, sigma = sgma)
  }
)
```

```{r}
betaE <- sapply(SimD1summ, function(x) x$beta)
STe <- sapply(SimD1summ, function(x) x$St)
sigmaE <- sapply(SimD1summ, function(x) x$sigma)
```

```{r}
betaEm <- apply(betaE, 1, mean)
betaEq <- apply(betaE, 1, function(x) quantile(x, c(0.5, 0.025, 0.975)))
ptE <- apply(
  betaE, 1,
  function(x) {
    prb <- mean(x > 0)
    2 * pmax(0.5 / ncol(betaE), pmin(prb, 1 - prb))
  }
)
cbind(betaEm, t(betaEq), ptE)
```

```{r}
d11E <- STe * sigmaE
rndE <- rbind(d11E, sigmaE)
rndEm <- rowMeans(rndE)
rndEq <- apply(rndE, 1, function(x) quantile(x, c(0.5, 0.025, 0.975)))
cbind(rndEm, t(rndEq))
```

```{r}
names(sigmaE) <- names(STe) <- NULL
parSimD1 <- rbind(betaE, ST1 = STe, sigma = sigmaE)
parSimD1t <- data.frame(t(parSimD1), check.names = FALSE)
parSimD1t
```

```{r}
parSimD1s <- subset(parSimD1t, select = -`(Intercept)`)
densityplot(~ value | variable,
  data = reshape::melt(parSimD1s),
  scales = list(relation = "free"),
  plot.points = FALSE
)
```

```{r}
vis.lm2 <- lm(lm2.form, data = armd)
```

```{r}
RLRTstat <- -2 * as.numeric(logLik(vis.lm2, REML = TRUE) - logLik(fm16.1mer))
RLRTstat
```

```{r}
0.6 * pchisq(RLRTstat, 1, lower.tail = FALSE)
```

```{r}
library(RLRsim)
exactRLRT(fm16.1mer)
```

```{r}
lm2sim <- simulate(vis.lm2, nsim = 100)
RLRTstatSim <- apply(
  lm2sim, 2,
  function(y) {
    dfAux <- within(armd, visual <- y)
    lm0 <- lm(formula(vis.lm2), data = dfAux)
    llik0 <- as.numeric(logLik(lm0, REML = TRUE))
    llikA <- as.numeric(logLik(refit(fm16.1mer, y)))
    # RLRTstat <- -2 * (llik0 - llikA)
  }
)
mean(RLRTstat <= RLRTstatSim)
```

```{r}
fm16.2mer <-
  lmer(visual ~ visual0 + time + treat.f + treat.f:time + (1 | subject) + (0 + time | subject),
    data = armd
  )
fm16.2mer
```

```{r}
summ <- summary(fm16.2mer)
coef(summ)
```

```{r}
unlist(VarCorr(fm16.2mer))
```

```{r}
sigma(fm16.2mer)
```

```{r}
fm16.2aux <- update(fm16.2mer, . ~ . - treat.f:time)
anova(fm16.2aux, fm16.2mer)
```

```{r}
RML0 <- logLik(fm16.1mer)
RMLa <- logLik(fm16.2mer)
(RLRTstat <- -2 * as.numeric(RML0 - RMLa))
```

```{r}
.5 * pchisq(RLRTstat, 1, lower.tail = FALSE) +
  5 * pchisq(RLRTstat, 2, lower.tail = FALSE)
```

```{r}
# library(RLRsim)
# maux <- lmer(visual ~ visual0 + time + treat.f + treat.f:time + (0 + time|subject),
#              data = armd)
# exactRLRT(m = mAux, m0 = fm16.1mer, mA = fm16.2mer)
```
