---
title: "note04"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# 4 Doing Data Analysis with the Multilevel Model for Change

## 4.1 Example: Changes in adolescent alcohol use

```{r}
library(tidyverse)
alcohol1_pp <- read_csv("data/alcohol1_pp.csv")
alcohol1_pp
```

여기서

```{r}
alcohol1_pp %>%
  summarize(
    mpeer = mean(peer) %>% round(digits = 3),
    mcoa = mean(coa) %>% round(digits = 3)
  )
```

```{r}
alcohol1_pp %>%
  filter(id %in% c(4, 14, 23, 32, 41, 56, 65, 82)) %>%
  ggplot(aes(age, alcuse)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point() +
  coord_cartesian(xlim = c(13, 17), ylim = c(-1, 4)) +
  facet_wrap(~id, ncol = 4) +
  theme(panel.grid = element_blank())
```

```{r}
set.seed(4)

alcohol1_pp %>%
  # group_by(id) %>%
  # nest() %>%
  # ungroup() %>%
  # sample_n(size = 32, replace = TRUE) %>%
  # unnest(data) %>%
  mutate(coa = ifelse(coa == 0, "coa = 0", "coa = 1")) %>%
  ggplot(aes(age, alcuse, group = id)) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, size = 1 / 4) +
  coord_cartesian(xlim = c(13, 17), ylim = c(-1, 4)) +
  facet_wrap(~coa) +
  theme(panel.grid = element_blank())
```

```{r}
set.seed(4)

alcohol1_pp %>%
  # group_by(id) %>%
  # nest() %>%
  # ungroup() %>%
  # sample_n(size = 32, replace = T) %>%
  # unnest(data) %>%
  mutate(hp = ifelse(peer < mean(peer), "low peer", "high peer")) %>%
  mutate(hp = factor(hp, levels = c("low peer", "high peer"))) %>%
  ggplot(aes(x = age, y = alcuse, group = id)) +
  stat_smooth(method = "lm", se = F, size = 1 / 4) +
  coord_cartesian(
    xlim = c(13, 17),
    ylim = c(-1, 4)
  ) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~hp)
```

## 4.2 The composite specification of the multilevel model for change

## 4.3 Methods of estimation, revisited

## 4.4 First steps: Fitting two unconditional multilevel models for change

```{r}
library(brms)
```

```{r}
get_prior(
  data = alcohol1_pp,
  family = gaussian,
  alcuse ~ 1 + (1 | id)
)
```

```{r}
fit4.1 <-
  brm(
    data = alcohol1_pp,
    family = gaussian,
    alcuse ~ 1 + (1 | id),
    prior = c(
      prior(student_t(3, 1, 2.5), class = Intercept),
      prior(student_t(3, 0, 2.5), class = sd),
      prior(student_t(3, 0, 2.5), class = sigma)
    ),
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4,
    file = "fits/fit04.01"
  )
```

```{r}
fit4.1
```

```{r}
post <- posterior_samples(fit4.1)
glimpse(post[, 1:12])
```

```{r}
v <-
  post %>%
  select(sigma, sd_id__Intercept) %>%
  mutate(
    sigma_2_epsilon = sigma^2,
    sigma_2_0 = sd_id__Intercept^2
  )
v
```

```{r}
v %>%
  pivot_longer(everything()) %>%
  ggplot(aes(value)) +
  geom_vline(xintercept = c(.25, .5, .75, 1), color = "white") +
  geom_density(size = 0, fill = "black") +
  scale_x_continuous(
    NULL,
    limits = c(0, 1.25), breaks = seq(0, 1.25, by = .25)
  ) +
  scale_y_continuous(NULL, breaks = NULL) +
  facet_wrap(~name, scales = "free_y") +
  theme(panel.grid = element_blank())
```

```{r}
v %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarize(
    mean = mean(value),
    median = median(value),
    sd = sd(value),
    ll = quantile(value, prob = .025),
    ul = quantile(value, prob = .975)
  ) %>%
  mutate(across(where(is.double), round, digits = 3))
```

```{r}
v %>%
  transmute(rho = sd_id__Intercept^2 / (sd_id__Intercept^2 + sigma^2)) %>%
  ggplot(aes(rho)) +
  geom_density(size = 0, fill = "black") +
  scale_x_continuous(expression(rho), limits = 0:1) +
  theme(panel.grid = element_blank())
```

```{r}
v %>%
  transmute(rho = sd_id__Intercept^2 / (sd_id__Intercept^2 + sigma^2)) %>%
  summarise(
    mean = mean(rho),
    median = median(rho),
    sd = sd(rho),
    ll = quantile(rho, prob = .025),
    ul = quantile(rho, prob = .975)
  ) %>%
  mutate_if(is.double, round, digits = 3)
```

```{r}
fixef(fit4.1)
```

```{r}
post[1:6, 1:3]
```

```{r}
post %>%
  transmute(gamma_00_squared = b_Intercept^2) %>%
  summarise(
    mean = mean(gamma_00_squared),
    median = median(gamma_00_squared),
    sd = sd(gamma_00_squared),
    ll = quantile(gamma_00_squared, prob = .025),
    ul = quantile(gamma_00_squared, prob = .975)
  ) %>%
  mutate_if(is.double, round, digits = 3) %>%
  pivot_longer(everything())
```

```{r}
tibble(x = seq(-20, 20, length.out = 1e3)) %>%
  mutate(density = metRology::dt.scaled(x, df = 3, mean = 1, sd = 2.5)) %>%
  ggplot(aes(x, density)) +
  geom_vline(xintercept = 1, color = "white") +
  geom_line() +
  labs(
    title = expression(paste("prior for ", gamma[0][0])),
    x = "parameter space"
  ) +
  theme(panel.grid = element_blank())
```

```{r}
tibble(x = seq(from = 0, to = 20, length.out = 1e3)) %>%
  mutate(density = metRology::dt.scaled(x, df = 3, mean = 0, sd = 2.5)) %>%
  ggplot(aes(x = x, y = density)) +
  geom_vline(xintercept = 0, color = "white") +
  geom_line() +
  labs(
    title = expression(paste("prior for both ", sigma[0], " and ", sigma[epsilon])),
    x = "parameter space"
  ) +
  theme(panel.grid = element_blank())
```

```{r}
get_prior(
  data = alcohol1_pp,
  family = gaussian,
  alcuse ~ 0 + Intercept + age_14 + (1 + age_14 | id)
)
```

```{r}
fit4.2 <-
  brm(
    data = alcohol1_pp,
    family = gaussian,
    alcuse ~ 0 + Intercept + age_14 + (1 + age_14 | id),
    prior = c(
      prior(student_t(3, 0, 2.5), class = sd),
      prior(student_t(3, 0, 2.5), class = sigma),
      prior(lkj(1), class = cor)
    ),
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4,
    file = "fits/fit04.02"
  )
```

```{r}
print(fit4.2, digits = 3)
```

```{r}
conditional_effects(fit4.2)
conditional_effects(fit4.2, robust = FALSE)
conditional_effects(fit4.2, robust = FALSE, prob = .8)
```

```{r}
post <- posterior_samples(fit4.2)
v <- post %>%
  transmute(
    sigma_2_epsilon = sigma^2,
    sigma_2_0 = sd_id__Intercept^2,
    sigma_2_1 = sd_id__age_14^2,
    sigma_01 = sd_id__Intercept * cor_id__Intercept__age_14 * sd_id__age_14
  )
v
```

```{r}
v %>%
  pivot_longer(everything()) %>%
  ggplot(aes(value)) +
  geom_density(size = 0, fill = "black") +
  scale_y_continuous(NULL, breaks = NULL) +
  facet_wrap(~name, scales = "free") +
  theme(panel.grid = element_blank())
```

```{r}
v %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarise(
    mean = mean(value),
    median = median(value),
    sd = sd(value),
    ll = quantile(value, prob = .025),
    ul = quantile(value, prob = .975)
  ) %>%
  mutate(across(where(is.double), round, digits = 3))
```

```{r}
vc <- VarCorr(fit4.2, summary = FALSE)
str(vc)
```

```{r}
cbind(
  VarCorr(fit4.1, summary = F)[[2]][[1]],
  VarCorr(fit4.2, summary = F)[[2]][[1]]
) %>%
  data.frame() %>%
  mutate(across(everything(), ~ .^2)) %>%
  set_names(str_c("fit4.", 1:2)) %>%
  mutate(`fit4.1 - fit4.2` = fit4.1 - fit4.2) %>%
  pivot_longer(everything()) %>%
  mutate(name = factor(name, levels = c("fit4.1", "fit4.2", "fit4.1 - fit4.2"))) %>%
  ggplot(aes(value)) +
  geom_vline(xintercept = .5, color = "white") +
  geom_density(fill = "grey25", color = "transparent") +
  scale_y_continuous(NULL, breaks = NULL) +
  facet_wrap(~name, scales = "free_y", ncol = 3) +
  theme(panel.grid = element_blank())
```

```{r}
cbind(
  VarCorr(fit4.1, summary = F)[[2]][[1]],
  VarCorr(fit4.2, summary = F)[[2]][[1]]
) %>%
  data.frame() %>%
  mutate_all(~ .^2) %>%
  set_names(str_c("fit4.", 1:2)) %>%
  mutate(proportion_decline = (fit4.1 - fit4.2) / fit4.1) %>%
  summarise(
    mean = mean(proportion_decline),
    median = median(proportion_decline),
    sd = sd(proportion_decline),
    ll = quantile(proportion_decline, prob = .025),
    ul = quantile(proportion_decline, prob = .975)
  ) %>%
  mutate_if(is.double, round, digits = 3)
```

```{r}
post %>%
  ggplot(aes(x = cor_id__Intercept__age_14)) +
  geom_vline(xintercept = 0, color = "white") +
  geom_density(fill = "grey25", color = "transparent") +
  scale_x_continuous(expression(rho[0][1]), limits = c(-1, 1)) +
  scale_y_continuous(NULL, breaks = NULL) +
  theme(panel.grid = element_blank())
```

```{r}
post %>%
  transmute(sigma_2_residual_j = sd_id__Intercept^2 +
    sd_id__age_14^2 * 0 +
    2 * sd_id__Intercept * cor_id__Intercept__age_14 * sd_id__age_14 * 0 + sigma^2) %>%
  head()
```

```{r}
make_s2rj <- function(x) {
  post %>%
    transmute(sigma_2_residual_j = sd_id__Intercept^2 + sd_id__age_14^2 * x + 2 * sd_id__Intercept * cor_id__Intercept__age_14 * sd_id__age_14 * x + sigma^2) %>%
    pull()
}
```

```{r}
tibble(age = 14:16) %>%
  mutate(age_c = age - 14) %>%
  mutate(s2rj = map(age_c, make_s2rj)) %>%
  unnest(s2rj) %>%
  mutate(label = str_c("age = ", age)) %>%
  ggplot(aes(s2rj)) +
  geom_density(fill = "grey25", color = "transparent") +
  geom_vline(xintercept = 1, color = "grey92", linetype = 2) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "Behold the shape of longitudinal heteroscedasticity.",
    x = expression(sigma[italic(Residual[j])]^2)
  ) +
  facet_wrap(~label, scales = "free_y", ncol = 1) +
  theme(panel.grid = element_blank())
```

```{r}
make_rho_rj_rjp <- function(j, jp) {
  s2rj_j <- make_s2rj(j)
  s2rj_jp <- make_s2rj(jp)
  post %>%
    transmute(
      r = (sd_id__Intercept^2 +
        sd_id__Intercept * cor_id__Intercept__age_14 * sd_id__age_14 * (j + jp) +
        sd_id__age_14^2 * j * jp) /
        sqrt(s2rj_j * s2rj_jp)
    ) %>%
    pull()
}
```

```{r}
make_rho_rj_rjp(0, 1) %>% median()
make_rho_rj_rjp(1, 2) %>% median()
make_rho_rj_rjp(0, 2) %>% median()
```

```{r}
tibble(occasion = 1:3) %>%
  mutate(
    age_c = occasion - 1,
    j = c(1, 2, 1) - 1,
    jp = c(2, 3, 3) - 1
  ) %>%
  mutate(r = map2(j, jp, make_rho_rj_rjp)) %>%
  unnest(r) %>%
  mutate(label = str_c("occasions ", j + 1, " and ", jp + 1)) %>%
  ggplot(aes(r)) +
  geom_vline(xintercept = c(.25, .5, .75), color = "white") +
  geom_density(fill = "grey25", color = "transparent") +
  scale_x_continuous(expression(rho[Residual[italic(j)]][Residual[italic(j * minute)]]), limits = 0:1) +
  scale_y_continuous(NULL, breaks = NULL) +
  facet_wrap(~label, scales = "free_y", ncol = 1) +
  theme(panel.grid = element_blank())
```

```{r}
fitted(fit4.2, re_formula = NA) %>% head()
```

```{r}
fitted(fit4.2) %>% head()
```

```{r}
f <- fitted(fit4.2, summary = FALSE, re_formula = NA) %>%
  as_tibble() %>%
  set_names(1:ncol(.)) %>%
  rownames_to_column("iter")
f
```

```{r}
f <-
  f %>%
  pivot_longer(-iter, names_to = "row", values_to = "fitted") %>%
  mutate(row = as.integer(row)) %>%
  left_join(
    alcohol1_pp %>%
      mutate(row = 1:n()) %>%
      select(row, alcuse),
    by = "row"
  )
f
```

```{r}
f %>%
  summarize(
    r = cor(fitted, alcuse),
    r2 = r^2
  )
```

```{r}
f %>%
  mutate(iter = as.double(iter)) %>%
  group_by(iter) %>%
  summarize(
    r = cor(fitted, alcuse),
    r2 = cor(fitted, alcuse)^2
  )
```

```{r}
tibble(age_14 = 0:2) %>%
  mutate(fitted = map(age_14, ~ post$b_Intercept + post$b_age_14 * .)) %>%
  full_join(alcohol1_pp %>% select(id, age_14, alcuse),
    by = "age_14"
  ) %>%
  mutate(row = 1:n()) %>%
  unnest(fitted) %>%
  mutate(iter = rep(1:4000, times = alcohol1_pp %>% nrow())) %>%
  group_by(iter) %>%
  summarise(
    r = cor(fitted, alcuse),
    r2 = cor(fitted, alcuse)^2
  )
```

```{r}
cbind(
  VarCorr(fit4.1, summary = F)[[2]][[1]],
  VarCorr(fit4.2, summary = F)[[2]][[1]]
) %>%
  data.frame() %>%
  mutate_all(~ .^2) %>%
  set_names(str_c("fit4.", 1:2)) %>%
  mutate(r_2_epsilon = (fit4.1 - fit4.2) / fit4.1) %>%
  summarise(
    mean = mean(r_2_epsilon),
    median = median(r_2_epsilon),
    sd = sd(r_2_epsilon),
    ll = quantile(r_2_epsilon, prob = .025),
    ul = quantile(r_2_epsilon, prob = .975)
  ) %>%
  mutate_if(is.double, round, digits = 3)
```

```{r}
cbind(
  VarCorr(fit4.1, summary = F)[[2]][[1]],
  VarCorr(fit4.2, summary = F)[[2]][[1]]
) %>%
  data.frame() %>%
  mutate_all(~ .^2) %>%
  set_names(str_c("fit4.", 1:2)) %>%
  mutate(r_2_epsilon = (fit4.1 - fit4.2) / fit4.1) %>%
  ggplot(aes(x = r_2_epsilon)) +
  geom_vline(xintercept = 0, color = "white") +
  geom_density(fill = "grey25", color = "transparent") +
  scale_x_continuous(expression(Pseudo ~ italic(R)[epsilon]^2), limits = c(-1, 1)) +
  scale_y_continuous(NULL, breaks = NULL) +
  theme(panel.grid = element_blank())
```

## 4.5 Practical data analytic strategies for model building

```{r}
fit4.3 <-
  brm(
    data = alcohol1_pp,
    family = gaussian,
    alcuse ~ 0 + Intercept + age_14 + coa + age_14:coa + (1 + age_14 | id),
    prior = c(
      prior(student_t(3, 0, 2.5), class = sd),
      prior(student_t(3, 0, 2.5), class = sigma),
      prior(lkj(1), class = cor)
    ),
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4,
    file = "fits/fit04.03"
  )
```

```{r}
fit4.3
```

```{r}
conditional_effects(fit4.3)
```

```{r}
glimpse(fit4.3$data)
```

```{r}
fit4.4 <-
  update(fit4.3,
    newdata = alcohol1_pp %>% mutate(coa = factor(coa)),
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4,
    file = "fits/fit04.04"
  )
```

```{r}
conditional_effects(fit4.4, plot = FALSE)
```

```{r}
posterior_summary(fit4.3)[1:4, ] %>% round(digits = 3)
posterior_summary(fit4.4)[1:4, ] %>% round(digits = 3)
```

```{r}
ce <- conditional_effects(fit4.4, effects = "age_14:coa")
str(ce)
```

```{r}
plot(ce)
```

```{r}
ce %>%
  plot(
    points = T,
    point_args = list(size = 1 / 4, alpha = 1 / 4, width = .05, height = .05, color = "black"),
    theme = theme(panel.grid = element_blank(), aspect.ratio = 1)
  )
```

```{r}
plot(ce,
  theme = theme(
    legend.position = "none",
    panel.grid = element_blank(),
    aspect.ratio = 1
  ),
  plot = FALSE
)[[1]] +
  annotate(
    geom = "text",
    x = 2.1, y = c(.95, 1.55),
    label = str_c("coa = ", 0:1),
    hjust = 0, size = 3.5
  ) +
  scale_fill_brewer(type = "qual") +
  scale_color_brewer(type = "qual") +
  scale_x_continuous("age", limits = c(-1, 3), labels = 13:17) +
  scale_y_continuous(limits = c(0, 2), breaks = 0:2)
```

```{r}
VarCorr(fit4.2)[[2]]
```

```{r}
posterior_summary(fit4.2)["sigma", ]
```

```{r}
bind_cols(
  posterior_samples(fit4.2) %>%
    transmute(
      fit2_sigma_2_0 = sd_id__Intercept^2,
      fit2_sigma_2_1 = sd_id__age_14^2
    ),
  posterior_samples(fit4.3) %>%
    transmute(
      fit3_sigma_2_0 = sd_id__Intercept^2,
      fit3_sigma_2_1 = sd_id__age_14^2
    )
) %>%
  mutate(
    `decline~'in'~sigma[0]^2` = (fit2_sigma_2_0 - fit3_sigma_2_0) / fit2_sigma_2_0,
    `decline~'in'~sigma[1]^2` = (fit2_sigma_2_1 - fit3_sigma_2_1) / fit2_sigma_2_1
  ) %>%
  pivot_longer(contains("decline")) %>%
  ggplot(aes(x = value)) +
  geom_vline(xintercept = 0, color = "white") +
  geom_density(fill = "grey25", color = "transparent") +
  scale_x_continuous(NULL, limits = c(-5, 2)) +
  scale_y_continuous(NULL, breaks = NULL) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~name, labeller = label_parsed, ncol = 1)
```

```{r}
bind_cols(
  posterior_samples(fit4.2) %>%
    transmute(
      fit2_sigma_2_0 = sd_id__Intercept^2,
      fit2_sigma_2_1 = sd_id__age_14^2
    ),
  posterior_samples(fit4.3) %>%
    transmute(
      fit3_sigma_2_0 = sd_id__Intercept^2,
      fit3_sigma_2_1 = sd_id__age_14^2
    )
) %>%
  mutate(
    `decline~'in'~sigma[0]^2` = (fit2_sigma_2_0 - fit3_sigma_2_0) / fit2_sigma_2_0,
    `decline~'in'~sigma[1]^2` = (fit2_sigma_2_1 - fit3_sigma_2_1) / fit2_sigma_2_1
  ) %>%
  pivot_longer(contains("decline")) %>%
  group_by(name) %>%
  summarise(
    mean = mean(value),
    median = median(value),
    sd = sd(value),
    ll = quantile(value, prob = .025),
    ul = quantile(value, prob = .975)
  )
```

```{r}
bind_cols(
  posterior_samples(fit4.2) %>%
    transmute(
      fit2_sigma_2_0 = sd_id__Intercept^2,
      fit2_sigma_2_1 = sd_id__age_14^2
    ),
  posterior_samples(fit4.3) %>%
    transmute(
      fit3_sigma_2_0 = sd_id__Intercept^2,
      fit3_sigma_2_1 = sd_id__age_14^2
    )
) %>%
  summarize(across(everything(), median)) %>%
  transmute(
    `% decline in sigma_2_0` = 100 * (fit2_sigma_2_0 - fit3_sigma_2_0) / fit2_sigma_2_0,
    `% decline in sigma_2_1` = 100 * (fit2_sigma_2_1 - fit3_sigma_2_1) / fit2_sigma_2_1
  )
```

```{r}
fit4.5 <-
  brm(
    data = alcohol1_pp,
    family = gaussian,
    alcuse ~ 0 + Intercept + age_14 + coa + peer + age_14:coa + age_14:peer +
      (1 + age_14 | id),
    prior = c(
      prior(student_t(3, 0, 2.5), class = sd),
      prior(student_t(3, 0, 2.5), class = sigma),
      prior(lkj(1), class = cor)
    ),
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4,
    file = "fits/fit04.05"
  )
```

```{r}
print(fit4.5, digits = 3)
```

```{r}
v <- posterior_samples(fit4.5) %>%
  transmute(
    sigma_2_epsilon = sigma^2,
    sigma_2_0 = sd_id__Intercept^2,
    sigma_2_1 = sd_id__age_14^2,
    sigma_01 = sd_id__Intercept * cor_id__Intercept__age_14 * sd_id__age_14
  )
v
```

```{r}
v %>%
  pivot_longer(everything()) %>%
  ggplot(aes(value)) +
  geom_density(size = 0, fill = "black") +
  scale_y_continuous(NULL, breaks = NULL) +
  facet_wrap(~name, scales = "free") +
  theme(panel.grid = element_blank())
```

```{r}
v %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarize(mean = mean(value))
```

```{r}
v %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarize(
    mean = mean(value),
    median = median(value),
    sd = sd(value),
    ll = quantile(value, prob = .025),
    ul = quantile(value, prob = .975)
  ) %>%
  mutate(across(where(is.double), round, digits = 3))
```

```{r}
bind_cols(
  posterior_samples(fit4.2) %>%
    transmute(
      fit4.2_sigma_2_epsilon = sigma^2,
      fit4.2_sigma_2_0 = sd_id__Intercept^2,
      fit4.2_sigma_2_1 = sd_id__age_14^2
    ),
  posterior_samples(fit4.5) %>%
    transmute(
      fit4.5_sigma_2_epsilon = sigma^2,
      fit4.5_sigma_2_0 = sd_id__Intercept^2,
      fit4.5_sigma_2_1 = sd_id__age_14^2
    )
) %>%
  summarise_all(median) %>%
  mutate(
    `% decline in sigma_2_epsilon` = 100 * (fit4.2_sigma_2_epsilon - fit4.5_sigma_2_epsilon) / fit4.2_sigma_2_epsilon,
    `% decline in sigma_2_0` = 100 * (fit4.2_sigma_2_0 - fit4.5_sigma_2_0) / fit4.2_sigma_2_0,
    `% decline in sigma_2_1` = 100 * (fit4.2_sigma_2_1 - fit4.5_sigma_2_1) / fit4.2_sigma_2_1
  ) %>%
  pivot_longer(contains("%")) %>%
  select(name, value)
```

```{r}
fit4.6 <-
  brm(
    data = alcohol1_pp,
    family = gaussian,
    alcuse ~ 0 + Intercept + age_14 + coa + peer + age_14:peer + (1 + age_14 | id),
    prior = c(
      prior(student_t(3, 0, 2.5), class = sd),
      prior(student_t(3, 0, 2.5), class = sigma),
      prior(lkj(1), class = cor)
    ),
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4,
    file = "fits/fit04.06"
  )
```

```{r}
prior_summary(fit4.6)
```

```{r}
prior_summary(fit4.6)
```

```{r}
print(fit4.6, digits = 3)
```

```{r}
v <-
  as_draws_df(fit4.6) %>%
  transmute(
    sigma_2_epsilon = sigma^2,
    sigma_2_0 = sd_id__Intercept^2,
    sigma_2_1 = sd_id__age_14^2,
    sigma_01 = sd_id__Intercept * cor_id__Intercept__age_14 * sd_id__age_14
  )
v %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarize(
    mean = mean(value),
    median = median(value),
    sd = sd(value),
    ll = quantile(value, prob = .025),
    ul = quantile(value, prob = .975)
  ) %>%
  mutate(across(where(is.double), round, digits = 3))
```

```{r}
post <- as_draws_df(fit4.3)
post %>%
  select(starts_with("b_"))
```

```{r}
post %>%
  select(starts_with("b_")) %>%
  transmute(
    pi_0_coa0 = b_Intercept + b_coa * 0,
    pi_1_coa0 = b_age_14 + `b_age_14:coa` * 0,
    pi_0_coa1 = b_Intercept + b_coa * 1,
    pi_1_coa1 = b_age_14 + `b_age_14:coa` * 1
  ) %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarize(posterior_mean = mean(value) %>% round(digits = 3))
```

```{r}
post %>%
  select(starts_with("b_")) %>%
  mutate(y = b_Intercept + b_coa * 1 + b_age_14 * 0 + `b_age_14:coa` * 0 * 1) # %>%
# summarize(pi_0_coa1 = mean(y))
```

```{r}
nd <-
  tibble(age_14 = seq(0, 2, length.out = 30))
f <- fitted(fit4.2, newdata = nd, re_formula = NA) %>%
  data.frame() %>%
  bind_cols(nd) %>%
  mutate(age = age_14 + 14)
f
```

```{r}
f %>%
  ggplot(aes(x = age)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = "grey75", alpha = 3 / 4) +
  geom_line(aes(y = Estimate)) +
  scale_y_continuous("alcuse", breaks = 0:2, limits = c(0, 2)) +
  xlim(13, 17) +
  theme(panel.grid = element_blank()) +
  theme(aspect.ratio = 1)
```

```{r}
nd <- crossing(
  coa = 0:1,
  peer = c(.655, 1.381)
) %>%
  expand(nesting(coa, peer),
    age_14 = seq(0, 2, length = 30)
  )
nd
```

```{r}
f <-
  fitted(fit4.6,
    newdata = nd,
    re_formula = NA
  ) %>%
  data.frame() %>%
  bind_cols(nd) %>%
  mutate(
    age = age_14 + 14,
    coa = if_else(coa == 0, "coa = 0", "coa = 1"),
    peer = factor(peer)
  )
glimpse(f)
```

```{r}
f %>%
  ggplot(aes(age, color = peer, fill = peer)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), size = 0, alpha = 1 / 4) +
  geom_line(aes(y = Estimate, size = peer)) +
  scale_size_manual(values = c(1 / 2, 1)) +
  scale_fill_manual(values = c("blue3", "red3")) +
  scale_color_manual(values = c("blue3", "red3")) +
  scale_y_continuous("alcuse", breaks = 0:2) +
  labs(subtitle = "High peer values are in red; low ones are in blue.") +
  coord_cartesian(xlim = c(13, 17)) +
  facet_wrap(~coa) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )
```

```{r}
alcohol1_pp %>%
  glimpse()
```

```{r}
fit4.7 <-
  update(
    fit4.6,
    newdata = alcohol1_pp,
    alcuse ~ 0 + Intercept + age_14 + coa + cpeer + age_14:cpeer + (1 + age_14 | id),
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4,
    file = "fits/fit04.07"
  )
```

```{r}
fit4.8 <-
  update(
    fit4.6,
    newdata = alcohol1_pp,
    alcuse ~ 0 + Intercept + age_14 + ccoa + peer + age_14:peer + (1 + age_14 | id),
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4,
    file = "fits/fit04.08"
  )
```

```{r}
fixef(fit4.6) %>% round(digits = 3)
fixef(fit4.7) %>% round(digits = 3)
fixef(fit4.8) %>% round(digits = 3)
```

## 4.6 Comparing models using deviance statistics

```{r}
log_lik(fit4.1) %>%
  glimpse()
```

```{r}
ll <-
  log_lik(fit4.1) %>%
  data.frame() %>%
  mutate(ll = rowSums(.)) %>%
  mutate(deviance = -2 * ll) %>%
  relocate(ll, deviance)
ll
```

```{r}
ll %>%
  ggplot(aes(deviance)) +
  geom_density(fill = "grey25", size = 0) +
  scale_y_continuous(NULL, breaks = NULL) +
  theme(panel.grid = element_blank())
```

```{r}
waic(fit4.1)
```

```{r}
fit4.1 <- add_criterion(fit4.1, "waic")
fit4.1$criteria$waic
```

```{r}
l_fit4.1 <- loo(fit4.1)
l_fit4.1
```

```{r}
loo::pareto_k_table(l_fit4.1)
```

```{r}
plot(l_fit4.1)
```

```{r}
library(loo)
pareto_k_ids(l_fit4.1, threshold = .7)
```

```{r}
alcohol1_pp[c(33, 79, 177), ]
```

```{r}
l_fit4.1$diagnostics$pareto_k
```

```{r}
fit4.1 <- add_criterion(fit4.1, c("loo", "waic"))
fit4.2 <- add_criterion(fit4.2, c("loo", "waic"))
fit4.3 <- add_criterion(fit4.3, c("loo", "waic"))
fit4.5 <- add_criterion(fit4.5, c("loo", "waic"))
```

```{r}
ws <- loo_compare(fit4.1, fit4.2, fit4.3, fit4.5, criterion = "waic")
ws
```

```{r}
cbind(
  waic_diff = ws[, 1] * -2,
  se = ws[, 2] * 2
)
```

```{r}
print(ws, simplify = FALSE)
```

```{r}
loo_compare(fit4.1, fit4.2, fit4.3, fit4.5, criterion = "loo") %>%
  print(simplify = F)
```

## 4.7 Using Wald statistics to test composite hypotheses about fixed effects

```{r}
mu_peer <- mean(alcohol1_pp$peer)
nd <-
  tibble(
    coa = 0,
    peer = mu_peer,
    age_14 = seq(0, 2, length.out = 30)
  )
nd
```

```{r}
f <-
  fitted(fit4.6,
    newdata = nd,
    re_formula = NA
  ) %>%
  data.frame() %>%
  bind_cols(nd) %>%
  mutate(age = age_14 + 14)
f %>%
  ggplot(aes(age)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), size = 0, alpha = 1 / 4) +
  geom_line(aes(y = Estimate)) +
  scale_y_continuous("alcuse", breaks = 0:2, limits = c(0, 2)) +
  labs(subtitle = "Zero is credible for neither\nthe intercept nor the slope.") +
  coord_cartesian(xlim = c(13, 17)) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    aspect.ratio = 1
  )
```

```{r}
f %>%
  select(age, Estimate, Q2.5, Q97.5) %>%
  filter(age %in% c(14, 16)) %>%
  mutate(across(where(is.double), round, digits = 2))
```

```{r}
nd <- tibble(
  coa = 0,
  peer = mu_peer,
  age_14 = 0:1
)
f <-
  fitted(fit4.6,
    newdata = nd,
    re_formula = NA,
    summary = FALSE
  ) %>%
  data.frame()
f
```

```{r}
f %>%
  transmute(difference = X2 - X1) %>%
  ggplot(aes(difference)) +
  geom_density(size = 0, fill = "grey25") +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    subtitle = "Based on 4,000 posterior draws, not a single one\nsuggests the slope is even close to zero. Rather, the\nposterior mass is concentrated around 0.25.",
    x = expression(paste(gamma[0][1], " (i.e., the difference between the two time points)"))
  ) +
  coord_cartesian(xlim = 0:1) +
  theme(panel.grid = element_blank(), )
```

```{r}
f %>%
  mutate(difference = X2 - X1) %>%
  summarize(
    mean = mean(difference),
    median = median(difference),
    ll = quantile(difference, probs = .025),
    ul = quantile(difference, probs = .975)
  )
```

```{r}
nd <-
  crossing(
    age_14 = c(0, 2),
    cpeer = c(-.363, .363)
  ) %>%
  mutate(coa = 0)
nd
```

```{r}
f <-
  fitted(fit4.7,
    newdata = nd,
    re_formula = NA,
    summary = FALSE
  ) %>%
  data.frame() %>%
  transmute(
    `difference at 14` = X2 - X1,
    `difference at 16` = X4 - X3,
    `difference in differences` = `difference at 16` - `difference at 14`
  )
f
```

```{r}
f %>%
  pivot_longer(everything()) %>%
  ggplot(aes(value)) +
  geom_density(size = 0, fill = "grey25") +
  scale_y_continuous(NULL, breaks = NULL) +
  xlab("different differences") +
  facet_wrap(~name, scales = "free_y") +
  theme(panel.grid = element_blank())
```

```{r}
f %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarise(
    mean = mean(value),
    median = median(value),
    ll = quantile(value, probs = .025),
    ul = quantile(value, probs = .975)
  ) %>%
  mutate(across(where(is.double), round, digits = 3))
```

## 4.8 Evaluating the tenability of a model's assumptions

```{r}
library(broom)
o <-
  alcohol1_pp %>%
  nest(data = c(-id, -coa, -peer)) %>%
  mutate(ols = map(data, ~ lm(data = ., alcuse ~ 1 + age_14))) %>%
  mutate(tidy = map(ols, tidy)) %>%
  unnest(tidy) %>%
  mutate(term = factor(term,
    levels = c("(Intercept)", "age_14"),
    labels = c("pi[0]", "pi[1]")
  ))
o
```

```{r}
o %>%
  select(coa:peer, term:estimate) %>%
  pivot_longer(coa:peer) %>%
  ggplot(aes(value, estimate)) +
  geom_hline(yintercept = 0, color = "white") +
  geom_point(alpha = 2 / 3) +
  facet_grid(term ~ name, scale = "free", labeller = label_parsed) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(size = 11)
  )
```

```{r}
o %>%
  select(coa:peer, term:estimate) %>%
  pivot_longer(coa:peer) %>%
  group_by(term, name) %>%
  nest() %>%
  mutate(r = map_dbl(data, ~ cor(.)[2, 1] %>% round(digits = 2)))
```

```{r}
e <-
  residuals(fit4.6) %>%
  data.frame()
e
```

```{r}
e %>%
  ggplot(aes(sample = Estimate)) +
  geom_hline(yintercept = 0, color = "white") +
  geom_qq() +
  ylim(-2, 2) +
  labs(
    x = "Normal score",
    y = expression(hat(epsilon)[italic(ij)])
  ) +
  theme(panel.grid = element_blank())
```

```{r}
e %>%
  bind_cols(alcohol1_pp %>% select(id)) %>%
  mutate(z = (Estimate - mean(Estimate)) / sd(Estimate)) %>%
  ggplot(aes(id, z)) +
  geom_hline(yintercept = 0, color = "white") +
  geom_point() +
  scale_y_continuous(
    expression(italic(std) ~ hat(epsilon)[italic(ij)]),
    limits = c(-3.5, 3.5)
  ) +
  theme(panel.grid = element_blank())
```

```{r}
z <- ranef(fit4.6)
str(z)
```

```{r}
z[[1]][1:6, , "Intercept"]
z[[1]][1:6, , "age_14"]
```

```{r}
z <-
  rbind(
    z[[1]][, , "Intercept"],
    z[[1]][, , "age_14"]
  ) %>%
  data.frame() %>%
  mutate(ranef = rep(c("hat(zeta)[0][italic(i)]", "hat(zeta)[1][italic(i)]"),
    each = n() / 2
  ))
glimpse(z)
```

```{r}
z %>%
  ggplot(aes(sample = Estimate)) +
  geom_hline(yintercept = 0, color = "white") +
  geom_qq() +
  ylim(-1, 1) +
  labs(
    x = "Normal score",
    y = NULL
  ) +
  facet_wrap(~ranef, labeller = label_parsed, nrow = 1) +
  theme(panel.grid = element_blank())
```

```{r}
as_tibble(z) %>%
  bind_cols(
    bind_rows(
      alcohol1_pp %>% distinct(id),
      alcohol1_pp %>% distinct(id)
    )
  ) %>%
  mutate(ranef = str_c("italic(std)~", ranef)) %>%
  group_by(ranef) %>%
  mutate(z = (Estimate - mean(Estimate)) / sd(Estimate)) %>%
  ggplot(aes(id, z)) +
  geom_hline(yintercept = 0, color = "white") +
  geom_point() +
  scale_y_continuous(NULL, limits = c(-3, 3)) +
  facet_wrap(~ranef, labeller = label_parsed, nrow = 1) +
  theme(panel.grid = element_blank())
```

```{r}
z %>%
  bind_cols(
    bind_rows(
      alcohol1_pp %>% distinct(id),
      alcohol1_pp %>% distinct(id)
    )
  ) %>%
  ggplot(aes(id, Estimate, ymin = Q2.5, ymax = Q97.5)) +
  geom_hline(yintercept = 0, color = "white") +
  geom_pointrange(shape = 20, size = 1 / 3) +
  ylab(NULL) +
  facet_wrap(~ranef, labeller = label_parsed, nrow = 1) +
  theme(panel.grid = element_blank())
```

```{r}
e %>%
  bind_cols(alcohol1_pp) %>%
  ggplot(aes(age, Estimate)) +
  geom_hline(yintercept = 0, color = "white") +
  geom_point(alpha = 1 / 4) +
  ylab(expression(hat(epsilon)[italic(ij)])) +
  coord_cartesian(xlim = c(13, 17), ylim = c(-2, 2)) +
  theme(panel.grid = element_blank())
```

```{r}
z %>%
  bind_cols(
    bind_rows(
      alcohol1_pp %>% distinct(id, coa, peer),
      alcohol1_pp %>% distinct(id, coa, peer)
    )
  ) %>%
  select(Estimate, ranef, coa, peer) %>%
  pivot_longer(-c(Estimate, ranef)) %>%
  ggplot(aes(value, Estimate)) +
  geom_hline(yintercept = 0, color = "white") +
  geom_point(alpha = 1 / 3) +
  ylim(-1, 1) +
  labs(
    x = "covariate value",
    y = NULL
  ) +
  facet_grid(ranef ~ name, labeller = label_parsed, scales = "free") +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(size = 10)
  )
```

## 4.9 Model-based (Empirical Bayes) estimates of the individual growth parameters

```{r}
alcohol1_pp %>%
  select(id:coa, cpeer, alcuse) %>%
  filter(id == 23)
```

```{r}
post_23 <-
  as_draws_df(fit4.7) %>%
  select(starts_with("b_")) %>%
  mutate(
    `pi[0][",23"]` = b_Intercept + b_coa * 1 + b_cpeer * -1.018,
    `pi[1][",23"]` = b_age_14 + `b_age_14:cpeer` * -1.018
  )
post_23
```

```{r}
post_23 %>%
  pivot_longer(starts_with("pi")) %>%
  group_by(name) %>%
  summarize(
    mean = mean(value),
    ll = quantile(value, probs = .025),
    ul = quantile(value, probs = .975)
  ) %>%
  mutate(across(where(is.double), round, digits = 3))
```

```{r}
post_23 %>%
  pivot_longer(starts_with("pi")) %>%
  ggplot(aes(value)) +
  geom_density(size = 0, fill = "grey25") +
  xlab("participant-specific parameter estimates") +
  facet_wrap(~name, labeller = label_parsed, scales = "free_y") +
  theme(panel.grid = element_blank())
```

```{r}
post_23 <-
  posterior_samples(fit4.7) %>%
  select(starts_with("b_"), contains("23"))
post_23
```

```{r}
post_23 <-
  post_23 %>%
  mutate(
    `pi[0][",23"]` = b_Intercept + b_coa * 1 + b_cpeer * -1.018 + `r_id[23,Intercept]`,
    `pi[1][",23"]` = b_age_14 + `b_age_14:cpeer` * -1.018 + `r_id[23,age_14]`
  )
post_23
```

```{r}
post_23 %>%
  pivot_longer(starts_with("pi")) %>%
  group_by(name) %>%
  summarize(
    mean = mean(value),
    ll = quantile(value, probs = .025),
    ul = quantile(value, probs = .975)
  ) %>%
  mutate(across(where(is.double), round, digits = 3))
```

```{r}
post_23 %>%
  pivot_longer(starts_with("pi")) %>%
  ggplot(aes(value)) +
  geom_density(size = 0, fill = "grey25") +
  scale_y_continuous(NULL, breaks = NULL) +
  xlab("participant-specific parameter estimates") +
  theme(panel.grid = element_blank()) +
  facet_wrap(~name, labeller = label_parsed, scales = "free_y")
```

```{r}
nd <-
  alcohol1_pp %>%
  select(id:coa, age_14:alcuse, cpeer) %>%
  filter(id %in% c(4, 14, 23, 32, 41, 56, 65, 82)) %>%
  mutate(id_label = ifelse(id < 10, str_c("0", id), id)) %>%
  mutate(id_label = str_c("id = ", id_label))
nd
```

```{r}
f <-
  fitted(fit4.7,
    newdata = nd
  ) %>%
  data.frame() %>%
  bind_cols(nd)
f
```

```{r}
f %>%
  ggplot(aes(age, Estimate)) +
  geom_line(size = 1) +
  scale_y_continuous("alcuse", breaks = 0:4, limits = c(-1, 4)) +
  xlim(13, 17) +
  facet_wrap(~id_label, ncol = 4) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )
```

```{r}
f_gamma_only <-
  fitted(fit4.7,
    newdata = nd,
    re_formula = NA
  ) %>%
  data.frame() %>%
  bind_cols(nd)
f_gamma_only
```

```{r}
f %>%
  ggplot(aes(age)) +
  geom_line(aes(y = Estimate), size = 1) +
  geom_line(
    data = f_gamma_only,
    aes(y = Estimate),
    size = 1 / 2
  ) +
  stat_smooth(
    data = nd,
    aes(y = alcuse),
    method = "lm", se = F,
    color = "black", linetype = 2, size = 1 / 2
  ) +
  geom_point(
    data = nd,
    aes(y = alcuse)
  ) +
  scale_y_continuous("alcuse", breaks = 0:4, limits = c(-1, 4)) +
  xlim(13, 17) +
  facet_wrap(~id_label, ncol = 4)
```

```{r}
f %>%
  ggplot(aes(age, Estimate)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = "grey75") +
  geom_line(size = 1) +
  geom_point(data = nd, aes(y = alcuse)) +
  scale_y_continuous("alcuse", breaks = 0:4, limits = c(-1, 4)) +
  facet_wrap(~id_label, ncol = 4)
```

```{r}
nd_expanded <-
  alcohol1_pp %>%
  select(id, coa, cpeer) %>%
  filter(id %in% c(4, 14, 23, 32, 41, 56, 65, 82)) %>%
  expand(nesting(id, coa, cpeer),
    age_14 = seq(0, 2, length.out = 30)
  ) %>%
  mutate(
    id_label = ifelse(id < 10, str_c("0", id), id),
    id_label = str_c("id = ", id_label),
    age = age_14 + 14
  )
f <- fitted(fit4.7,
  newdata = nd_expanded
) %>%
  data.frame() %>%
  bind_cols(nd_expanded)
f
```

```{r}
f %>%
  ggplot(aes(age, Estimate)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = "grey75") +
  geom_line(size = 1) +
  geom_point(data = nd, aes(y = alcuse)) +
  scale_y_continuous("alcuse", breaks = 0:4, limits = c(-1, 4)) +
  xlim(13, 17) +
  facet_wrap(~id_label, ncol = 4) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )
```
