---
title: "note03"
output: html_notebook
---

# 3 Introducing the Multilevel Model for Change

## 3.1 What is the purpose of the multilevel model for change?

```{r}
library(tidyverse)
```

```{r}
early_int <-
  tibble(
    id = rep(c(68, 70:72, 902, 904, 906, 908), each = 3),
    age = rep(c(1, 1.5, 2), times = 8),
    cog = c(
      103, 119, 96, 106, 107, 96, 112, 86, 73, 100, 93, 87,
      119, 93, 99, 112, 98, 79, 89, 66, 81, 117, 90, 76
    ),
    program = rep(1:0, each = 12)
  )

early_int
```

```{r}
early_int <- early_int %>%
  mutate(age_c = age - 1)
early_int
```

```{r}
load("data/early_int_sim.rda")
```

```{r}
early_int_sim
```

## 3.2 The level-1 submodel for individual change

```{r}
early_int %>%
  ggplot(aes(x = age, y = cog)) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  geom_point() +
  scale_x_continuous(breaks = c(1, 1.5, 2)) +
  ylim(50, 150) +
  facet_wrap(~id, ncol = 4) +
  theme(panel.grid = element_blank())
```

```{r}
d <- tibble(
  id = "i",
  age = c(1, 1.5, 2),
  cog = c(95, 100, 135)
)
d
```

```{r}
fit3.1 <-
  lm(data = d, cog ~ age)
summary(fit3.1)
```

```{r}
f <-
  fitted(fit3.1) %>%
  data.frame() %>%
  set_names("fitted") %>%
  bind_cols(d)
f
```

```{r}
path <-
  tibble(
    age = c(1, 2, 2),
    cog = c(90, 90, 130)
  )
text <-
  tibble(
    tibble(
      age = c(1.2, 1.65, 2.15, 1.125, 2.075),
      cog = c(105, 101, 137, 75, 110),
      label = c(
        "epsilon[italic(i)][1]", "epsilon[italic(i)][2]",
        "epsilon[italic(i)][3]", "pi[0][italic(i)]", "pi[1][italic(i)]"
      )
    )
  )

arrow <-
  tibble(
    age = c(1.15, 1.6, 2.1, 1.1),
    xend = c(1.01, 1.51, 2.01, 1.01),
    cog = c(103, 101, 137, 78),
    yend = c(92.5, 105, 132.5, 89)
  )

f %>%
  ggplot(aes(age, cog)) +
  geom_point() +
  geom_line(aes(y = fitted)) +
  geom_path(data = path, linetype = 2, size = 1 / 2) +
  geom_segment(aes(xend = age, y = cog, yend = fitted), linetype = 3, size = 1 / 4) +
  geom_segment(
    data = arrow,
    aes(xend = xend, yend = yend),
    arrow = arrow(length = unit(0.1, "cm")), size = 1 / 4
  ) +
  geom_text(
    data = text,
    aes(label = label),
    size = c(4, 4, 4, 5, 5), parse = T
  ) +
  annotate(
    geom = "text",
    x = 1.5, y = 85, label = "1 year"
  ) +
  scale_x_continuous(breaks = c(1, 1.5, 2)) +
  ylim(50, 150) +
  theme(panel.grid = element_blank()) +
  theme(aspect.ratio = 1)
```

```{r}
early_int_sim %>%
  ggplot(aes(age, cog)) +
  geom_smooth(aes(group = id), method = "lm", se = FALSE, size = 1 / 6) +
  geom_smooth(method = "lm", se = FALSE, size = 2) +
  scale_x_continuous(breaks = c(1, 1.5, 2)) +
  ylim(50, 150) +
  theme(panel.grid = element_blank()) +
  theme(aspect.ratio = 1)
```

```{r}
by_id <-
  early_int_sim %>%
  mutate(age_c = age - 1) %>%
  group_by(id) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(data = ., cog ~ age_c)))
by_id
```

```{r}
library(broom)

by_id <-
  by_id %>%
  mutate(
    tidy = map(model, tidy),
    glance = map(model, glance)
  )
by_id
```

```{r}
by_id %>%
  unnest(tidy) %>%
  filter(term == "(Intercept)") %>%
  pull(estimate) %>%
  stem()
```

```{r}
unnest(by_id, tidy) %>%
  filter(term == "age_c") %>%
  pull(estimate) %>%
  stem()
```

```{r}
by_id %>%
  unnest(glance) %>%
  pull(sigma) %>%
  `^`(2) %>%
  stem(scale = 1)
```

## 3.3 The level-2 submodel for systematic interindividual differences in change

```{r}
early_int_sim <-
  early_int_sim %>%
  mutate(label = str_c("program = ", program))

early_int_sim %>%
  ggplot(aes(age, cog, color = label)) +
  geom_smooth(aes(group = id), method = "lm", se = FALSE, size = 1 / 6) +
  geom_smooth(method = "lm", se = FALSE, size = 2) +
  scale_x_continuous(breaks = c(1, 1.5, 2)) +
  ylim(50, 150) +
  facet_wrap(~label) +
  scale_color_viridis_d(option = "B", begin = .33, end = .67) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )
```

```{r}
n <- 31
nd <- tibble(age_c = seq(0, 1, length.out = n))

p <-
  by_id %>%
  mutate(fitted = map(model, ~ predict(., newdata = nd))) %>%
  unnest(fitted) %>%
  mutate(
    age = seq(1, 2, length.out = n),
    program = ifelse(id < 900, 1, 0)
  ) %>%
  group_by(program, age) %>%
  summarize(
    min = min(fitted),
    max = max(fitted)
  ) %>%
  mutate(label = str_c("program = ", program))
p
```

```{r}
text <-
  tibble(
    age = 1.01,
    cog = c(101.5, 110),
    label = c("program = 0", "program = 1"),
    text = c("Average population trajectory,", "Average population trajectory,"),
    angle = c(345.7, 349)
  )

math <-
  tibble(
    age = 1.01,
    cog = c(94.5, 103),
    label = c("program = 0", "program = 1"),
    text = c("gamma[0][0] + gamma[10](italic(age) - 1)", "(gamma[0][0] + gamma[10]) + (gamma[10] + gamma[11]) (italic(age) - 1)"),
    angle = c(345.7, 349)
  )

p %>%
  ggplot(aes(x = age)) +
  geom_ribbon(aes(ymin = min, ymax = max, fill = label), alpha = 1 / 3) +
  stat_smooth(
    data = early_int_sim,
    aes(y = cog, color = label), method = "lm", se = FALSE, size = 1
  ) +
  geom_text(
    data = text,
    aes(y = cog, label = text, angle = angle),
    hjust = 0
  ) +
  geom_text(
    data = math,
    aes(y = cog, label = text, angle = angle),
    hjust = 0, parse = TRUE
  ) +
  scale_x_continuous(breaks = c(1, 1.5, 2)) +
  scale_y_continuous(limits = c(50, 150)) +
  facet_wrap(~label) +
  scale_fill_viridis_d(option = "B", begin = .33, end = .67) +
  scale_color_viridis_d(option = "B", begin = .33, end = .67) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )
```

## 3.4 Fitting the multilevel model for change to data

## 3.5 Examining estimated fixed effects

```{r}
library(brms)
```

```{r}
fit3.2 <-
  brm(
    data = early_int_sim,
    family = gaussian,
    formula = cog ~ 0 + Intercept + age_c + program + age_c:program + (1 + age_c | id),
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    control = list(adapt_delta = 0.95),
    seed = 3,
    file = "fits/fit03.02"
  )
```

```{r}
fit3.2
```

```{r}
fit3.2_ <-
  update(fit3.2,
    formula = cog ~ 1 + age_c + program + age_c:program + (1 + age_c | id),
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    control = list(adapt_delta = 0.95),
    seed = 3,
    file = "fits/fit03.02_"
  )
```

```{r}
fit3.2_
```

```{r}
fixef(fit3.2) %>% round(digits = 2)
```

```{r}
fixefs <- fixef(fit3.2)
fixefs[1, 1] + fixefs[3, 1]
```

```{r}
fixefs[2, 1] + fixefs[4, 1]
```

```{r}
fixefs[1, 1] + fixefs[3, 1] * 0
```

```{r}
fixefs[2, 1] + fixefs[4, 1] * 0 * 1
```

```{r}
fixefs[1, 1] + fixefs[3, 1] * 1
```

```{r}
fixefs[2, 1] + fixefs[4, 1] * 1 * 1
```

```{r}
crossing(age_c = 0:1, program = 0:1) %>%
  mutate(
    cog = fixefs[1, 1] + fixefs[2, 1] * age_c + fixefs[3, 1] * program +
      fixefs[4, 1] * age_c * program,
    age = age_c + 1,
    size = ifelse(program == 1, 1 / 5, 3),
    program = factor(program, levels = c("0", "1"))
  ) %>%
  ggplot(aes(age, cog, group = program)) +
  geom_line(aes(size = program)) +
  scale_size_manual(values = c(1, 1 / 2)) +
  scale_x_continuous(breaks = c(1, 1.5, 2)) +
  ylim(50, 150) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    aspect.ratio = 1
  )
```

```{r}
post <- posterior_samples(fit3.2)
glimpse(post)
```

```{r}
post %>%
  select(starts_with("b_")) %>%
  head()
```

```{r}
post %>%
  pivot_longer(starts_with("b_")) %>%
  ggplot(aes(value)) +
  geom_density(color = "transparent", fill = "gray25") +
  scale_y_continuous(NULL, breaks = NULL) +
  facet_wrap(~name, scales = "free") +
  theme(panel.grid = element_blank())
```

```{r}
VarCorr(fit3.2)[[1]] %>% str()
```

```{r}
VarCorr(fit3.2)[[1]][[1]]
```

```{r}
VarCorr(fit3.2)[[1]][[2]]
```

```{r}
VarCorr(fit3.2)[[1]][[3]]
```

```{r}
posterior_samples(fit3.2) %>%
  mutate(
    `sigma[0]^2` = sd_id__Intercept^2,
    `sigma[1]^2` = sd_id__age_c^2,
    `sigma[0][1]` = sd_id__Intercept * cor_id__Intercept__age_c * sd_id__age_c,
    `sigma[epsilon]^2` = sigma^2
  ) %>%
  select(starts_with("sigma[")) %>%
  pivot_longer(everything(), values_to = "posterior") %>%
  ggplot(aes(posterior)) +
  geom_density(color = "transparent", fill = "grey33") +
  scale_y_continuous(NULL, breaks = NULL) +
  facet_wrap(~name, scales = "free", labeller = label_parsed) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(size = 12)
  )
```

## 3.6 Examining estimated variance components

## 3.7 Bonus: How did you simulate that data?

```{r}
# how many people are we simulating?
n <- 103 - 8

# what's the variance/covariance matrix?
sigma <- matrix(c(
  124.64, -36.41,
  -36.41, 12.29
),
ncol = 2
)

# what's our mean structure?
mu <- c(0, 0)

# set the seed and simulate!
set.seed(3)
z <-
  MASS::mvrnorm(n = n, mu = mu, Sigma = sigma) %>%
  data.frame() %>%
  set_names("zeta_0", "zeta_1")

z
```

```{r}
g <-
  tibble(
    id = 1:n,
    gamma_00 = 107.84,
    gamma_01 = 6.85,
    gamma_10 = -21.13,
    gamma_11 = 5.27
  )

g
```

```{r}
# set the seed for the second `mutate()` line
set.seed(3)

early_int_sim <-
  bind_cols(g, z) %>%
  mutate(program = rep(1:0, times = c(54, 41))) %>%
  expand(nesting(id, gamma_00, gamma_01, gamma_10, gamma_11, zeta_0, zeta_1, program),
    age_c = c(0, 0.5, 1)
  ) %>%
  mutate(epsilon = rnorm(n(), mean = 0, sd = sqrt(74.24))) %>%
  mutate(
    pi_0 = gamma_00 + gamma_01 * program + zeta_0,
    pi_1 = gamma_10 + gamma_11 * program + zeta_1
  ) %>%
  mutate(cog = pi_0 + pi_1 * age_c + epsilon)

early_int_sim
```

```{r}
early_int_sim <-
  early_int_sim %>%
  mutate(
    age = age_c + 1,
    cog = round(cog, digits = 0),
    id = ifelse(id > 54, id + 900, id)
  )

early_int_sim
```

```{r}
early_int_sim <-
  early_int_sim %>%
  select(id, age, cog, program, age_c) %>%
  full_join(early_int,
    by = c("id", "age", "cog", "program", "age_c")
  ) %>%
  arrange(id, age)

glimpse(early_int_sim)
```

```{r}
save(early_int_sim,
  file = "data/early_int_sim.rda"
)
```
